---
title: "Fig. 4 - Maifeld validation"
author: "Gelsomina Nicola Kaufhold"
date: "2025-06-18"
output: html_document
---

# Figure 4: Validation in independent studies: Maifeld


# Load packages
```{r message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(readxl)
library(rstatix)
library(dplyr)
library(microbiome)
library(coin)
library(rstatix)
library(caret)
library(stats)

boxplot.theme.legend <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black",angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "top")

boxplot.theme <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")
```


# Load and prepare data
```{r}
PS.CORONA <- readRDS("rowdata/PS.CORONA.rds")
# Visite should be used as factor
sample_data(PS.CORONA)$Visite <- as.factor(sample_data(PS.CORONA)$Case)
levels(sample_data(PS.CORONA)$Visite) <- c("1" = "baseline", "2" = "fasting", "3" = "follow-up")

meta.corona <- data.frame(sample_data(PS.CORONA))

meta.corona.red <- meta.corona %>%
  select(Patient, Visite, age, sex, BMI, BP_sphygm_syst, BP_sphygm_diast, HDL, LDL)

# calculate a variaty of diversity indices
tab <- microbiome::alpha(PS.CORONA, index = "all") 

# load meta data
Samp <- sample_data(PS.CORONA)

# merge with alpha diversity indicies
samp_alpha_corona <- (merge(Samp, tab, by = 0))

alpha.corona <- samp_alpha_corona %>%
  select(Row.names, Patient, diversity_shannon) %>%
  column_to_rownames(var = "Row.names")
```

## Defining responders
```{r}
meta.corona.baseline <- meta.corona %>%
  filter(Visite == "baseline") %>%
  rename(BMI.baseline = BMI)

meta.corona.recovery <- meta.corona %>%
  filter(Visite == "follow-up") %>%
  select(Patient, BMI) %>%
  rename(BMI.recovery = BMI)

meta.corona.bmi <- meta.corona.baseline %>%
  left_join(meta.corona.recovery, by = "Patient")

meta.corona.bmi$bmi.change <- meta.corona.bmi$BMI.baseline - meta.corona.bmi$BMI.recovery

obere.corona <- 0.8233315

meta.corona.bmi$responder.calculated <- ifelse(meta.corona.bmi$bmi.change > obere.corona, "1", "0")
```

## Prepare data for making a prediction

### Species abundance
```{r}
species <- tax_glom(physeq = PS.CORONA, taxrank = "Species")
tax_table <- data.frame(tax_table(species))

species <- tax_table %>%
  select(Species) %>%
  rownames_to_column(var = "motus")

abundance.corona <- data.frame(otu_table(PS.CORONA))

# assign the species names to the otus
abundance.species <- abundance.corona %>%
  rownames_to_column(var = "motus") %>%
  left_join(species, by = "motus") %>%
  replace_na(list(Species = "unassigned"))

abundance.species$motus <- NULL

# remove [....]
abundance.species$Species <- gsub("\\[.*?\\]", "", abundance.species$Species)
abundance.species$Species <- gsub("/.*", "", abundance.species$Species)
# ensure that there is no extra space at the end
abundance.species$Species <- trimws(abundance.species$Species)

# sum the same species
library(dplyr)
abundance_species_sum <- abundance.species %>%
  group_by(Species) %>%
  dplyr::summarize(across(starts_with("Sample"), ~ sum(.x, na.rm = TRUE))) %>%
  column_to_rownames(var = "Species") %>%
  filter(rowSums(.) > 0)# Keep only rows with a non-zero sum


species.corona <- abundance_species_sum

###############
##### call species of interest
###################
species <- data.frame(t(species.corona))
species_int <- c("NA.Faecalibacterium.sp..CAG.74.", "NA.Clostridium.species.incertae.sedis", "Faecalibacterium.prausnitzii.1.", "NA Roseburia species incertae sedis", "NA.Faecalibacterium.species.incertae.sedis", "Oscillibacter sp. 57_20", "Faecalibacterium.prausnitzii", "Akkermansia.muciniphila..CAG.154.UNK.MGS.1.54_46.Phil8.", "NA.Prevotella.species.incertae.sedis", "NA.Prevotellaceae.species.incertae.sedis")

species.int.corona <- c("NA.Faecalibacterium.species.incertae.sedis", "NA.Clostridiales.species.incertae.sedis", "Faecalibacterium.prausnitzii", "NA.Faecalibacterium.species.incertae.sedis", "NA.Roseburia.species.incertae.sedis", "Oscillibacter.sp..57_20", "Akkermansia.muciniphila", "NA.Prevotella.species.incertae.sedis", "NA.Bacteroides.rodentium")

bact_int <- species %>%
  select(species.int.corona) %>%
  rename("Oscillibacter.sp..57_20." = "Oscillibacter.sp..57_20", "NA.Bacteroides.rodentium.uniformis." = "NA.Bacteroides.rodentium", "Akkermansia.muciniphila..CAG.154.UNK.MGS.1.54_46.Phil8." = "Akkermansia.muciniphila")
```

### combine with meta data
```{r}
meta.corona.model <- meta.corona.red %>%
  rename(sys_mean = BP_sphygm_syst, bp_dia_mean = BP_sphygm_diast, hdl = HDL, ldl = LDL)

corona.test.model <- cbind(bact_int, meta.corona.model)
corona.baseline.model <- corona.test.model %>%
  filter(Visite == "baseline")


```

## make prediciton
```{r}
model_short <- readRDS("rowdata/model.heilig.rds")
predicted_corona <- predict(model_short, newdata = corona.baseline.model)
print(predicted_corona)

fit_b

predicted_corona <- predict(model.short.no.ldl, newdata = corona.baseline.model)
print(predicted_corona)



## add prediction to data
corona.baseline.model$responder.model <- predicted_corona

corona.baseline.model <- corona.baseline.model %>%
  mutate(responder.model =
           recode(responder.model, 
                  "NonResponder" = "non-responder",
                            "Responder"    = "responder"))

corona.responder <- corona.baseline.model %>%
  select(Patient, responder.model)




#####################################
responder.true <- meta.corona.bmi %>%
  select(Patient, responder.calculated) %>%
  rename(responder.true = responder.calculated)

corona.prediction <- corona.baseline.model %>%
  left_join(responder.true, by = "Patient")


corona.prediction$responder.true = factor(corona.prediction$responder.true, 
                                          levels = c(0, 1), 
                                          labels = c("non-responder", "responder"))


```

```{r}
library(pROC)
library(dplyr)

library(pROC)

# 1) get probabilities
probs <- predict(final_fit_no_ldl, newdata = corona.prediction, type = "prob")

# 2) add the responder probability to your dataset
corona.prediction$prob_responder <- probs$responder

# 3) make ROC and AUC
roc_obj <- roc(corona.prediction$responder.true,
               corona.prediction$prob_responder,
               levels = c("non-responder","responder"))

auc(roc_obj)
plot(roc_obj, main = "ROC curve")

```




## do we see higher bmi loss in respondern?
```{r}
meta.corona.new <- meta.corona.bmi %>%
  left_join(corona.responder, by = "Patient")

meta.corona.new$responder.model <- factor(meta.corona.new$responder.model, 
                                          levels = c("0", "1"), 
                                          labels = c("non-responder", "responder"))


meta.corona.new$responder.model <- factor(meta.corona.new$responder.model, 
                                          levels = c("non-responder", "responder"))

meta.corona.new$responder.calculated <- factor(meta.corona.new$responder.calculated, 
                                               levels = c(0, 1),
                                          labels = c("non-responder", "responder"))



meta.corona.new %>%
  ggplot(aes(x = responder.model, y = bmi.change, fill = responder.model)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  scale_fill_manual("", values = c("non-responder" ="#fc8d59", "responder" = "#998ec3")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  theme_classic() +
  boxplot.theme.legend +
  ylim(c(0, 6)) +
  labs(y = "Δ BMI") -> fig.4.BMI.change.corona
fig.4.BMI.change.corona
ggsave("fig.4.BMI.change.corona.pdf", plot = fig.4.BMI.change.corona, width = 5, height = 5)
#############
#### statistics
################
meta.corona.new %>% 
  wilcox_test(formula = bmi.change ~ responder.model, data = .)

meta.corona.new %>% 
  pairwise_wilcox_test(formula = bmi.change ~ responder.model,
                       paired = FALSE)

```

## BMI changes in CORONA
```{r}
##### add predicted responder to all data
meta.corona <- meta.corona %>%
  left_join(corona.responder, by = "Patient")

responder.calculated <-meta.corona.bmi %>%
  select(Patient, responder.calculated)

meta.corona <- meta.corona %>%
  left_join(responder.calculated, by = "Patient")

meta.corona$responder.model <- factor(meta.corona$responder.model, 
                                          levels = c(0, 1), 
                                          labels = c("non-responder", "responder"))

meta.corona$responder.calculated <- factor(meta.corona$responder.calculated, 
                                          levels = c(0, 1), 
                                          labels = c("non-responder", "responder"))



  
meta.corona %>%
  ggplot(aes(x = Visite, y = BMI, fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "BMI") -> fig.4.bmi.CORONA.responder
fig.4.bmi.CORONA.responder
#ggsave("fig.4.bmi.CORONA.responder.pdf", plot = fig.4.bmi.CORONA.responder, width = 5, height = 5)
```

## Features splittet for non responder / responder

### systolic Blood pressure
```{r}
meta.corona %>%
  ggplot(aes(x = Visite, y = BP_sphygm_syst, fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "sys BP") -> fig.4.sysBP.CORONA.responder
fig.4.sysBP.CORONA.responder

####################
##### statistics
#################
####### differences at baseline
meta.corona %>%
  filter(Visite == "baseline") %>%
  group_by(responder.model) %>%
  dplyr::summarize(mean = mean(BP_sphygm_syst),
            sd = sd(BP_sphygm_syst))

meta.corona %>% 
  filter(Visite == "baseline") %>%
  pairwise_wilcox_test(
    formula = BP_sphygm_syst ~ responder.model, 
    paired = FALSE)

######## changes over time
###### non-responder
meta.corona %>%
  filter(responder.model == "non-responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(BP_sphygm_syst),
            sd = sd(BP_sphygm_syst))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "non-responder") %>%
  pairwise_wilcox_test(
    formula = BP_sphygm_syst ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

###### responder
meta.corona %>%
  filter(responder.model == "responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(BP_sphygm_syst),
            sd = sd(BP_sphygm_syst))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "responder") %>%
  pairwise_wilcox_test(
    formula = BP_sphygm_syst ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")


```

### LDL
```{r}
meta.corona %>%
  ggplot(aes(x = Visite, y = LDL, fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "LDL") -> fig.4.LDL.CORONA.responder
fig.4.LDL.CORONA.responder

####################
##### statistics
#################
####### differences at baseline
meta.corona %>%
  filter(Visite == "baseline") %>%
  group_by(responder.model) %>%
  dplyr::summarize(mean = mean(LDL),
            sd = sd(LDL))

meta.corona %>% 
  filter(Visite == "baseline") %>%
  pairwise_wilcox_test(
    formula = LDL ~ responder.model, 
    paired = FALSE)

######## changes over time
###### non-responder
meta.corona %>%
  filter(responder.model == "non-responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(LDL),
            sd = sd(LDL))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "non-responder") %>%
  pairwise_wilcox_test(
    formula = LDL ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

###### responder
meta.corona %>%
  filter(responder.model == "responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(LDL),
            sd = sd(LDL))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "responder") %>%
  pairwise_wilcox_test(
    formula = LDL ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")
```

### Bacteria
```{r}
corona.bact <- corona.test.model %>%
  rownames_to_column(var = "Samp")

meta.corona <- meta.corona %>%
  left_join(corona.bact, by = c("Patient", "Visite"))

####################
######### Oscilli
##################
meta.corona %>%
  ggplot(aes(x = Visite, y = Oscillibacter.sp..57_20., fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Abundance") -> fig.4.Oscilli.CORONA.responder
fig.4.Oscilli.CORONA.responder

####################
##### statistics
#################
####### differences at baseline
meta.corona %>%
  filter(Visite == "baseline") %>%
  group_by(responder.model) %>%
  dplyr::summarize(mean = mean(Oscillibacter.sp..57_20.),
            sd = sd(Oscillibacter.sp..57_20.))

meta.corona %>% 
  filter(Visite == "baseline") %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ responder.model, 
    paired = FALSE)

######## changes over time
###### non-responder
meta.corona %>%
  filter(responder.model == "non-responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(Oscillibacter.sp..57_20.),
            sd = sd(Oscillibacter.sp..57_20.))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "non-responder") %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

###### responder
meta.corona %>%
  filter(responder.model == "responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(Oscillibacter.sp..57_20.),
            sd = sd(Oscillibacter.sp..57_20.))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "responder") %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")


##########################
########### Faecalibacterium
############################
meta.corona %>%
  ggplot(aes(x = Visite, y = NA.Faecalibacterium.species.incertae.sedis, fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Abundance") -> fig.4.Faecali.CORONA.responder
fig.4.Faecali.CORONA.responder

####################
##### statistics
#################
####### differences at baseline
meta.corona %>%
  filter(Visite == "baseline") %>%
  group_by(responder.model) %>%
  dplyr::summarize(mean = mean(NA.Faecalibacterium.species.incertae.sedis),
            sd = sd(NA.Faecalibacterium.species.incertae.sedis))

meta.corona %>% 
  filter(Visite == "baseline") %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ responder.model, 
    paired = FALSE)

######## changes over time
###### non-responder
meta.corona %>%
  filter(responder.model == "non-responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(NA.Faecalibacterium.species.incertae.sedis),
            sd = sd(NA.Faecalibacterium.species.incertae.sedis))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "non-responder") %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

###### responder
meta.corona %>%
  filter(responder.model == "responder") %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(NA.Faecalibacterium.species.incertae.sedis),
            sd = sd(NA.Faecalibacterium.species.incertae.sedis))

complete_meta <- meta.corona %>% 
  group_by(Patient) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  filter(responder.model == "responder") %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

```

### baseline only
```{r}
corona.baseline.model$responder <- factor(corona.baseline.model$responder,
                                          levels = c(0,1), 
                                          labels = c("non-responder", "responder"))
corona.baseline.model %>%
  filter(Visite == "baseline") %>%
  ggplot(aes(x = responder, y = NA.Faecalibacterium.species.incertae.sedis, fill = responder)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Abundance") -> fig.4.CORONA.faecali.baseline

 fig.4.CORONA.faecali.baseline
ggsave("fig.4.CORONA.faecali.baseline.pdf", plot =fig.4.CORONA.faecali.baseline, width = 3, height = 5)

corona.baseline.model %>%
  filter(Visite == "baseline") %>%
  ggplot(aes(x = responder, y = Oscillibacter.sp..57_20., fill = responder)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Abundance") -> fig.4.CORONA.Oscilli.baseline

fig.4.CORONA.Oscilli.baseline
ggsave("fig.4.CORONA.Oscilli.baseline.pdf", plot =fig.4.CORONA.Oscilli.baseline, width = 3, height = 5)

corona.baseline.model %>%
  filter(Visite == "baseline") %>%
  ggplot(aes(x = responder, y = sys_mean, fill = responder)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "sys BP (mmHg)") -> fig.4.CORONA.bp.baseline

fig.4.CORONA.bp.baseline
ggsave("fig.4.CORONA.bp.baseline.pdf", plot =fig.4.CORONA.bp.baseline, width = 3, height = 5)

corona.baseline.model %>%
  filter(Visite == "baseline") %>%
  ggplot(aes(x = responder, y = ldl, fill = responder)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "LDL (mg/dL)") -> fig.4.CORONA.ldl.baseline

fig.4.CORONA.ldl.baseline
ggsave("fig.4.CORONA.ldl.baseline.pdf", plot =fig.4.CORONA.ldl.baseline, width = 3, height = 5)

CORONA.stats <- pivot_longer(corona.baseline.model, cols = c(sys_mean, ldl, Oscillibacter.sp..57_20., NA.Faecalibacterium.species.incertae.sedis), names_to = "feature", values_to = "value")
CORONA.stats $feature <- as.factor(CORONA.stats$feature)
CORONA.stats %>%
  group_by(feature) %>%
  pairwise_wilcox_test(formula = value ~ responder, paired = FALSE) -> CORONA.stats

library(writexl)
#write_xlsx(CORONA.stats, "CORONA.stats.xlsx")
```


### Changes in BMI
```{r}
meta.corona %>%
  ggplot(aes(x = Visite, y = BMI.x, fill = responder.model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "BMI") -> corona.responder.model.bmi
corona.responder.model.bmi

##### how much percent weight loss in respondern / non respondern
bmi.baseline <- meta.corona %>%
  select(Patient, SampleID, BMI.x, Visite, responder.model, responder.calculated) %>%
  filter(Visite == "baseline") %>%
  rename("BMI.baseline" = BMI.x)

bmi.refeeding <- meta.corona %>%
  select(Patient, SampleID, BMI.x, Visite, responder.model, responder.calculated) %>%
  filter(Visite == "follow-up") %>%
  rename("BMI.refeeding" = BMI.x)

## join both cases
bmi.dif <- bmi.baseline %>%
  full_join(bmi.refeeding, by = "Patient") %>%
  mutate(bmi.loss = abs(BMI.refeeding-BMI.baseline)) %>%
  mutate(bmi.percent = (bmi.loss/BMI.baseline)*100)

## responder model
bmi.responder.model <- bmi.dif %>%
  filter(responder.model.x == "responder")## 8 have higher bmi loss then 5%

## responder calculated
bmi.responder.calculated <- bmi.dif %>%
  filter(responder.calculated.x == "responder") %>%
  filter(bmi.percent > 5)



####################
##### statistics
#################
####### differences at baseline
meta.corona %>%
  filter(Visite == "baseline") %>%
  group_by(responder.calculated) %>%
  dplyr::summarize(mean = mean(BMI.x),
            sd = sd(BMI.x))

### differences at follow-up
meta.corona %>%
  filter(Visite == "follow-up") %>%
  group_by(responder.calculated) %>%
  dplyr::summarize(mean = mean(BMI.x),
            sd = sd(BMI.x))


## statistical test
meta.corona %>%
  group_by(Visite) %>%
  pairwise_wilcox_test(BMI.x ~ responder.model, 
                       paired = FALSE, 
                       p.adjust.method = "BH")


```


## Test models performance

### confusion matrix
```{r}
library(caret)
meta.corona.new <- meta.corona.bmi %>%
  left_join(corona.responder, by = "Patient")

meta.corona.new$responder.model <- factor(meta.corona.new$responder.model, 
                                          levels = c(0, 1), 
                                          labels = c("non-responder", "responder"))

meta.corona.new$responder.calculated <- factor(meta.corona.new$responder.calculated, 
                                          levels = c(0, 1), 
                                          labels = c("non-responder", "responder"))



confusionMatrix(data = meta.corona.new$responder.model, reference =  meta.corona.new$responder.calculated, positive = "responder") -> confmat_CORONA
confmat_CORONA


############################
######## Plot confusion matrix
#############################
library(scales)

# Create confusion matrix
conf_mat <- matrix(c(6, 1, 3, 13), nrow = 2, byrow = FALSE)
rownames(conf_mat) <- c("non-responder", "responder")
colnames(conf_mat) <- c("non-responder", "responder")

# Convert matrix to data frame for plotting
conf_mat_df <- as.data.frame.table(as.table(conf_mat))

# Add color column based on whether the prediction is correct
conf_mat_df$Color <- ifelse(conf_mat_df$Var1 == rownames(conf_mat) & conf_mat_df$Var2 == colnames(conf_mat), "Correct",
                            ifelse(conf_mat_df$Freq == 0, "No Data", "Incorrect"))

# Plot heatmap
ggplot(conf_mat_df, aes(x = Var2, y = Var1, fill = Color)) +
  geom_tile(color = "black") +
  geom_text(aes(label = format(Freq, big.mark = ",")), color = "black", size = 8) +  # Format text with commas
  scale_fill_manual("Prediction", values = c("Correct" = "#5ab4ac", "Incorrect" = "#e9a3c9", "No Data" = "white")) +
  labs(x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(legend.title = element_text(size = 24),
    axis.text = element_text(size = 24, color = "black"),
    axis.title = element_text(size = 24, color = "black", face = "bold"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        panel.border = element_rect(color = "black", fill = NA,  size = 1),
        legend.text = element_text(color = "black", size = 24),
        legend.position = "top") -> fig.4.head.map.conf.matrix.CORONA 

fig.4.head.map.conf.matrix.CORONA
```

### ROC curve
```{r}
predicted_probs <- predict(final_fit, newdata = corona.baseline.model, type = "prob")


meta.corona.resp <- meta.corona.baseline %>%
  rownames_to_column(var ="sample") %>%
  left_join(corona.responder, by = "Patient") %>%
  #rename(responder.model = responder) %>%
  left_join(responder.calculated, by = "Patient")

predicted_probs <- predicted_probs %>%
  rownames_to_column(var = "sample")

combined_data <- meta.corona.resp %>%
  left_join(predicted_probs, by = "sample") %>%
  column_to_rownames(var = "sample")
  

# Create a prediction object
library(ROCR)
pred <- prediction(combined_data$`1`, combined_data$responder.calculated)

# Compute the performance measures
perf <- performance(pred, measure = "tpr", x.measure = "fpr")

# Extract the data for plotting
roc_data <- data.frame(
  FPR = perf@x.values[[1]],
  TPR = perf@y.values[[1]],
  Cutoff = perf@alpha.values[[1]]
)

# Calculate and print the AUC
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
print(paste("Area Under the ROC Curve:", round(auc, 2)))

# Plot the ROC curve

ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(size = 1, color = "steelblue") +
  geom_abline(linetype = "dashed", color = "red") +  # Add diagonal line
  annotate("text", x = 0.6, y = 0.2, label = paste("AUC =", round(auc, 2)), size = 5) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), name = "1 - specifity") +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), name = "sensitivity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black"),
        panel.border = element_rect(linewidth = 1, fill = NA)) -> fig.4.ROC.CORONA.no.confidence
fig.4.ROC.CORONA.no.confidence
ggsave("fig.4.ROC.CORONA.no.confidence.pdf", plot = fig.4.ROC.CORONA.no.confidence, width = 5, height = 4)
```

### ROC curve with confidence intervall
```{r}
library(pROC)
combined_data$responder.calculated <- factor(combined_data$responder.calculated, 
                                             levels = c(0, 1),
                                             labels = c("non-responder", "responder"))
roc_obj <- roc(response = combined_data$responder.calculated,
               predictor = combined_data$`1`,
               levels = c("non-responder", "responder"),  
               direction = "<",
               drop.intermediate = FALSE) 


ci_obj <- ci.se(roc_obj, specificities = seq(0, 1, by = 0.01))

# Extract the data for plotting
ci_df <- data.frame(
  specificity = as.numeric(rownames(ci_obj)),
  TPR         = ci_obj[, "50%"],      # median TPR at each specificity
  TPR_lower   = ci_obj[, "2.5%"],     # lower 95% CI
  TPR_upper   = ci_obj[, "97.5%"]     # upper 95% CI
)

# Convert specificity -> FPR for plotting
ci_df$FPR <- 1 - ci_df$specificity


library(ggplot2)
ggplot(ci_df, aes(x = FPR, y = TPR)) +
  geom_ribbon(aes(ymin = TPR_lower, ymax = TPR_upper),
              fill = "red", alpha = 0.2) +
  geom_line(color = "red", size = 1.2) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(limits = c(0, 1), name = "1 - Specificity") +
  scale_y_continuous(limits = c(0, 1), name = "Sensitivity") + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black"),
        panel.border = element_rect(linewidth = 1, fill = NA)) -> fig.4.ROC.CORONA.with.confintervall
 fig.4.ROC.CORONA.with.confintervall
```

### Correlation of selection features
```{r}
library(corrplot)
library(Hmisc)
corona.baseline.model

# select feature for correlation matrix
cor_features <- corona.baseline.model %>% 
  select(Oscillibacter.sp..57_20.,
         NA.Faecalibacterium.species.incertae.sedis,
         sys_mean,
         ldl,
         BMI)

cor_names <- cor_features
colnames(cor_features) <- c("Oscilli sp.\n57_20", 
           "Faecali sp.\nincertae sedis", 
           "Sys BP", 
           "LDL", 
           "BMI")

cor_test_result <- rcorr(as.matrix(cor_features, type = "spearman"))  


print(cor_test_result$r)  # Correlation matrix
print(cor_test_result$P) # Matrix with p valies
cor_test_result$P[is.na(cor_test_result$P)] <- 1 # Matrix of p-values

# Extract the correlation matrix (r) and p-value matrix (P)
cor_matrix <- cor_test_result$r  
p_matrix <- cor_test_result$P    
corrplot(cor_matrix, 
         type = "upper",         
         order = "hclust",       
         p.mat = p_matrix,       
         sig.level = 0.05,       
         insig = "blank") 

#svg("fig.4.corrplot.CORONA.significant.only.svg", width = 8, height = 8)
corrplot(cor_matrix, 
         type = "upper",         
         order = "hclust",       
         p.mat = p_matrix,       
         sig.level = 0.05,       
         insig = "blank",        
         tl.col = "black",       
         tl.srt = 45,            
         tl.cex = 0.8,           
         cl.cex = 0.8,           
         addCoef.col = "black",  
         number.cex = 0.7,       
         col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(100), 
         mar = c(0, 0, 1, 0))
#dev.off()

#svg("fig.4.corrplot.CORONA.svg", width = 8, height = 8)
corrplot(cor_matrix, type = "upper", order = "hclust", 
         tl.col = "black", tl.cex = 1.5, cl.cex = 1.3,
         number.cex = 1.5)
#dev.off()


###### square plot
library(ggcorrplot)
cor_matrix <- cor_test_result$r
p_matrix <- cor_test_result$P
p_matrix[is.na(p_matrix)] <- 1  # Set NA p-values to 1 (non-significant)

# Plot
fig.4.CORONA.heatmap <- ggcorrplot(cor_matrix,
           method = "square",        
           type = "upper",           
           lab = TRUE,               # Show correlation numbers
           lab_size = 6,             # Correlation number size (~24 pt)
           colors = c("#0075DC", "white", "#E64B35FF"),  
           tl.cex = 1.5,             # Axis label size (~24 pt)
           legend.title = "Corr",    # Optional: set a custom legend title
           ggtheme = theme_minimal(base_size = 24)) +  # Base text size 24
  theme(legend.text = element_text(size = 20),
        legend.title = element_text(size = 24),
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        plot.title = element_text(size = 24))

fig.4.CORONA.heatmap
#ggsave("fig.4.CORONA.heatmap.pdf", plot = fig.4.CORONA.heatmap, width = 8, height = 8)
```


# Rerun the model in the CORONA dataset
What are the important features here?
```{r}
species.corona
meta.corona

## species of interest
species <- data.frame(t(species.corona))
species_int <- c("NA.Faecalibacterium.sp..CAG.74.", "NA.Clostridium.species.incertae.sedis", "Faecalibacterium.prausnitzii.1.", "NA Roseburia species incertae sedis", "NA.Faecalibacterium.species.incertae.sedis", "Oscillibacter sp. 57_20", "Faecalibacterium.prausnitzii", "Akkermansia.muciniphila..CAG.154.UNK.MGS.1.54_46.Phil8.", "NA.Prevotella.species.incertae.sedis", "NA.Prevotellaceae.species.incertae.sedis")

species.int.corona <- c("NA.Faecalibacterium.species.incertae.sedis", "NA.Clostridiales.species.incertae.sedis", "Faecalibacterium.prausnitzii", "NA.Faecalibacterium.species.incertae.sedis", "NA.Roseburia.species.incertae.sedis", "Oscillibacter.sp..57_20", "Akkermansia.muciniphila", "NA.Prevotella.species.incertae.sedis", "NA.Bacteroides.rodentium")

## what are the 50 most abundant species?
bact <- data.frame(t(species.corona))
features <- bact
index.keep.1 <- features[, colSums(features)!=0] 
index.keep_t = t(index.keep.1)
index.keep.50 <- index.keep_t[names(sort(rowSums(index.keep_t), decreasing = TRUE)[1:50]), ]
index.keep.50 <- as.data.frame(t(index.keep.50))
top50 <- colnames(index.keep.50)
bacteria_top50 <- bact %>%
  select(species.int.corona, top50)

#### Bind to meta data
meta.int <- meta.corona %>%
  select(BMI, body_fat_ratio, HDL, LDL, LDL_HDL_ratio, triglycerol, Visite, responder.calculated)

class_data <- cbind(meta.int, bacteria_top50)

class_data <- class_data %>%
  filter(Visite == "baseline") %>%
  select(!Visite)
```

Train the model 
```{r}
## ===============================
## Random Forest: Repeated CV + Nested CV (caret)
## ===============================

library(tidyverse)
library(caret)
library(pROC)
library(randomForest)


#####################
####### Repeated 80/20 with inner CV
#############################
# Outer split: random 80% train / 20% test, repeated 100 times.
# Inner CV: 5-fold inside each 80% to pick the best mtry.
# What you get: a distribution of performance (median, IQR), showing stability/variance across many random splits.
# → Answers “Could our result be a lucky split?”


set.seed(42)

## ---- 0) DATA PREP (expects df with responder_mean + predictors) ----
# Ensure: (i) no IDs/text columns, (ii) two-class factor with valid names, (iii) positive class first (necessary for ROC)
df <- class_data
dat <- df %>%
  mutate(
    responder.calculated = factor(
      ifelse(as.character(responder.calculated) %in% c("1", 1, "responder"),
             "Responder", "NonResponder"),
      levels = c("Responder", "NonResponder")
    )
  ) %>%
  drop_na()

# number of predictors
p <- ncol(dat) - 1

##############################################
## ---- 1) Helper: robust varImp extractor (builds an Overall column)
#####################################
### after training the model, this function pulls variable importance from caret::varImp() and returns a table with feature, its importance and how often it came up (iter)

get_vi <- function(fit, iter_or_fold){
  vi_obj <- caret::varImp(fit, scale = FALSE)
  imp_raw <- tryCatch(vi_obj$importance, error = function(e) as.data.frame(vi_obj))
  imp_df <- as.data.frame(imp_raw)

  if(!"Overall" %in% colnames(imp_df)){
    if(ncol(imp_df) > 1){
      imp_df$Overall <- rowMeans(imp_df, na.rm = TRUE)
    } else {
      colnames(imp_df) <- "Overall"
    }
  }

  tibble(
    Feature    = rownames(imp_df),
    Importance = imp_df[ , "Overall", drop = TRUE],
    iter       = iter_or_fold
  )
}


#######################################
## ---- 2) Common training controls & grid ----
##########################################
# mtry is the RF hyperparameter (how many features are considered at each split)
# inner 5-fold CV = we try different mtry values and pick the one with the best ROC

rf_ntree <- 1000
rf_grid  <- expand.grid(mtry = seq(1, min(p, floor(sqrt(p)) + 2), by = 1))

ctrl_inner <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,  
  savePredictions = "final",
  allowParallel = TRUE
)

##########################
## A) Repeated 80/20 Monte-Carlo CV (FULL MODEL)
#########################
# we do the resampling for 100 times

B <- 100
metrics_mc  <- vector("list", B)
imp_mc_list <- vector("list", B)

for(b in seq_len(B)){
  # stratified 80/20 split
  tr_idx <- createDataPartition(dat$responder.calculated, p = 0.8, list = FALSE)
  tr <- dat[tr_idx, ]
  te <- dat[-tr_idx, ]

  # train on 80% with inner CV (optimize ROC)
  fit_b <- train(
    responder.calculated ~ .,
    data = tr,
    method = "rf",
    metric = "ROC",
    trControl = ctrl_inner,
    tuneGrid = rf_grid,
    ntree = rf_ntree,
    importance = TRUE
  )

  # predict probs on held-out 20%
  prob <- predict(fit_b, newdata = te, type = "prob")[, "Responder"]
  pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                 levels = c("Responder", "NonResponder"))

  # metrics (positive class = Responder)
  roc_obj <- pROC::roc(response = te$responder.calculated,
                       predictor = prob,
                       levels = c("NonResponder","Responder"),
                       direction = "<")
  cm <- caret::confusionMatrix(pred, te$responder.calculated, positive = "Responder")

  metrics_mc[[b]] <- tibble(
    iter      = b,
    AUC       = as.numeric(pROC::auc(roc_obj)),
    Accuracy  = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    Best_mtry = fit_b$bestTune$mtry
  )

  imp_mc_list[[b]] <- get_vi(fit_b, b)
}

### summary of the result, save AUC. Accuaracy, Sensitivity, Specificity and which mtry was chosen
mc_metrics     <- bind_rows(metrics_mc)
mc_importance  <- bind_rows(imp_mc_list)

## Summaries you can report (median/IQR over repeats)
## gives median and IQR of the key metrics across the 100 repeats → this shows typical performance and its spread.
mc_summary <- mc_metrics %>%
  summarise(
    AUC_median = median(AUC, na.rm=TRUE),
    AUC_IQR_L  = quantile(AUC, 0.25, na.rm=TRUE),
    AUC_IQR_U  = quantile(AUC, 0.75, na.rm=TRUE),
    Acc_median = median(Accuracy, na.rm=TRUE),
    Sens_median= median(Sensitivity, na.rm=TRUE),
    Spec_median= median(Specificity, na.rm=TRUE)
  )
print(mc_summary)


## Feature stability: Top-k frequency across repeats
# median and IQR of the key metrics across the 100 repeats → this shows typical performance and its spread.
k <- 5
mc_stability <- mc_importance %>%
  group_by(iter) %>%
  slice_max(order_by = Importance, n = k, with_ties = FALSE) %>%
  ungroup() %>%
  count(Feature, name = "TopK_hits") %>%
  arrange(desc(TopK_hits)) %>%
  mutate(TopK_freq = TopK_hits / B)
mc_stability %>% print(n = 30)

mc_stability

# Mean importance across repeats (helps ranking)
mc_mean_importance_short <- mc_importance %>%
  group_by(Feature) %>%
  summarise(MeanImportance = mean(Importance, na.rm=TRUE),
            SDImportance   = sd(Importance, na.rm=TRUE)) %>%
  arrange(desc(MeanImportance))

mc_mean_importance_short
```


## Is LDL alone associated with the outcome?
Fit an univariate logistic regression as a control

```{r}
glm.data <- combined_data
glm.data$responder.model <- NULL
glm.data$Responder <- NULL
glm.data$NonResponder <- NULL

glm.data <- glm.data %>%
  filter(Visite == "baseline")

glm.data$responder.calculated <- factor(glm.data$responder.calculated, 
                                        levels = c(0,1), 
                                        labels = c("non-responder", "responder"))

glm_ldl <- glm(responder.calculated ~ LDL, data = glm.data, family = binomial())
prob_ldl <- predict(glm_ldl, type = "response")
roc_ldl  <- roc(glm.data$responder.calculated, prob_ldl,
                levels=c("non-responder","responder"), direction="<")
auc_ldl  <- as.numeric(auc(roc_ldl))
cat("External univariate LDL AUC:", round(auc_ldl,3), "\n")


####### plot the ROC curve
# ROC curve for univariate LDL logistic regression
plot(roc_ldl, 
     col = "blue", lwd = 2,
     main = "ROC curve: LDL-only logistic regression (Maifeld dataset)")

# Add diagonal line = random chance
abline(a = 0, b = 1, lty = 2, col = "gray")

# Add AUC to the plot
auc_val <- auc(roc_ldl)
legend("bottomright", 
       legend = paste("AUC =", round(auc_val, 2)), 
       col = "blue", lwd = 2, bty = "n")

```







