---
title: "Fig.3 - Model"
author: "Gelsomina Nicola Kaufhold"
date: "2025-06-18"
output: html_document
---

# Figure 3: Random Forest Model
## Subset of patient with long-term response in bodyweight 

# Load required packages
```{r}
library(phyloseq)
library(tidyverse)
library(readxl)
library(rstatix)
library(dplyr)
library(microbiome)
```

# Load data
```{r}
PS.tree <- readRDS("rowdata/PS.tree.rds")
PS.all.visits <- readRDS("rowdata/PS_all_Visits.rds")

boxplot.theme.legend <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black",angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "top")

boxplot.theme <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")

boxplot.theme.2 <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")
```


## Prepare machine learning data
### Prepare functional data of scfa production
```{r}
kegg_modules <- read_csv("rowdata/functions//kegg_module_annotation.csv")
kegg_modules <- kegg_modules[ ,c(0:127)]

kegg_modules <- kegg_modules %>%
  rename("modules" = 1) %>%
  column_to_rownames(var = "modules")

col_names_new <- gsub(pattern = "preprocessed\\.|filtered_qc\\.pair", replacement = "", colnames(kegg_modules))
col_names_new <- gsub(pattern = "_", replacement = ".", col_names_new)
colnames(kegg_modules) <- col_names_new

kegg_modules <- kegg_modules %>%
  select(long, everything()) %>%
  mutate(across(starts_with("V"), as.numeric))

summarized_kegg <- kegg_modules %>%
  group_by(long) %>%
  summarize(across(starts_with("V"), ~sum(.)))

summarized_kegg$long[is.na(summarized_kegg$long)] <- "undefined"

sum_kegg <- summarized_kegg %>%
  column_to_rownames(var = "long")

######################
######## select functions related to scfa production
################
modules_propionate <- summarized_kegg %>%
  filter(str_detect(long, "propionate"))


modules_butyrate <- summarized_kegg %>%
  filter(str_detect(long, "butyrate"))

modules_acetate <- summarized_kegg %>%
  filter(str_detect(long, "acetate")) %>%
  filter(long != "lysine fermentation to acetate and butyrate (3")

modules_tryptophan <- summarized_kegg %>%
  filter(str_detect(long, "Tryptophan"))

modules_tryptophan_1 <- summarized_kegg %>%
  filter(str_detect(long, "tryptophan"))

modules_int <- rbind(modules_propionate, modules_butyrate, modules_acetate, modules_tryptophan, modules_tryptophan_1)
modules_int <- modules_int %>%
  column_to_rownames(var = "long")
```

### Prepare metadata
```{r}
meta_data <- data.frame(sample_data(PS.tree))
meta_int <- meta_data %>%
  select(Probanden_ID, BMI, Sex, Visite, sys_mean, bp_dia_mean, hdl, ldl)

modules_int <- modules_int %>%
  select(starts_with(c("V2", "V3", "V4")))

modules_int <- t(modules_int)

modules_int <- modules_int[row.names(modules_int) %in% row.names(meta_int), ]

class_data <- cbind(meta_int, modules_int)
```

### Prepare species of interest
```{r}
names <- readRDS("rowdata/names.rds")
colnames(names) <- gsub('_', '.', colnames(names))
names <-names[, colnames(names) %in% row.names(meta_data) ]



species_int <- c("NA.Faecalibacterium.sp..CAG.74.", "NA.Clostridium.species.incertae.sedis", "Faecalibacterium.prausnitzii.1.", "NA.Roseburia.species.incertae.sedis", "NA.Faecalibacterium.species.incertae.sedis", "Oscillibacter.sp..57_20.", "Faecalibacterium.prausnitzii", "Akkermansia.muciniphila..CAG.154.UNK.MGS.1.54_46.Phil8.", "NA.Prevotella.species.incertae.sedis", "NA.Prevotellaceae.species.incertae.sedis")

bact_int <- data.frame(t(names))
bact_int <- bact_int %>%
  select(species_int)

## what are the 50 most abundant species?
bact <- data.frame(t(names))
features <- bact
index.keep.1 <- features[, colSums(features)!=0] 
index.keep_t = t(index.keep.1)
index.keep.50 <- index.keep_t[names(sort(rowSums(index.keep_t), decreasing = TRUE)[1:50]), ]
index.keep.50 <- as.data.frame(t(index.keep.50))
top50 <- colnames(index.keep.50)
bacteria_top50 <- bact %>%
  select(species_int, top50)

#### Bind to meta data
class_data <- cbind(meta_int, modules_int, bact_int, bacteria_top50)
```

### Create machine learning data
```{r}
tab <- microbiome::alpha(PS.tree, index = "all") 
alpha <- tab %>%
  select(diversity_shannon)

class_data_all <- cbind(meta_int, modules_int, bact_int, alpha)
```

## Cluster patients into responder and non responder according to bmi loss
```{r}
###########################
####### what is the normal bmi fluctuation?
##########################

meta.all <- data.frame(meta(PS.all.visits))

# include BMI into data
library(readxl)
hight <- read_excel("rowdata/weight.xlsx", 
    col_names = FALSE)

hight <- hight[ , c(7, 11)]
colnames(hight) <- c("proband_id", "Hight")

meta.all <- meta.all %>%
  left_join(hight, by = "proband_id") 

meta.all.bmi <- meta.all %>%
  dplyr::mutate(Hight = Hight / 100) %>%
  mutate(BMI = koerpergewicht / (Hight^2))

rownames(meta.all.bmi) <- meta.all.bmi$Novogene_ID

meta.all.bmi$BMI <- as.numeric(meta.all.bmi$BMI)

##################################
##### Fluctuation between V1 and V2
################################

V1.bmi <- meta.all.bmi %>%
  filter(Visite == "V1") %>%
  select(proband_id, BMI)

mean(V1.bmi$BMI)
sd(V1.bmi$BMI)

V2.bmi <- meta.all.bmi %>%
  filter(Visite == "V2") %>%
  select(proband_id, BMI)

bmi <- V1.bmi %>%
  full_join(V2.bmi, by = "proband_id")

bmi$dif <- (bmi$BMI.x - bmi$BMI.y)

bmi.mean <- bmi %>%
  drop_na(dif)

bmi.mean.red <- bmi.mean %>%
  filter(proband_id != "LEANER-19")

mean(bmi.mean.red$dif)
sd(bmi.mean.red$dif)


untere.mean <-  mean(bmi.mean$dif) -(2*sd(bmi.mean.red$dif))
obere.mean <- mean(bmi.mean.red$dif)+(2*sd(bmi.mean.red$dif))
# everyone with a BMI change above 0.56 is considered as a responder

##############################################
###### clustering patients according to their bmi loss
#####################################
samp.meta <- data.frame(sample_data(PS.tree))
samp.bmi <- samp.meta %>%
  select(Probanden_ID, Visite, BMI)

bmi.baseline <- samp.bmi %>%
  filter(Visite == "baseline 1") %>%
  select(Probanden_ID, BMI) %>%
  rename(bmi.baseline = BMI)

bmi.recovery <- samp.bmi %>%
  filter(Visite == "follow-up") %>%
  select(Probanden_ID, BMI) %>%
  rename(bmi.recovery = BMI)

bmi.total <- bmi.baseline %>%
  full_join(bmi.recovery, by = "Probanden_ID") %>%
  filter(!is.na(bmi.recovery))

bmi.total$dif <- bmi.total$bmi.baseline - bmi.total$bmi.recovery

bmi.total$responder_mean <- ifelse(test = bmi.total$dif > obere.mean, yes = 1, no = 0)

bmi.cluster.mean <- bmi.total %>%
  select(Probanden_ID, responder_mean)

class_data_all <- data.frame(class_data_all)
class_data_baseline <- class_data_all %>%
  filter(Visite == "baseline 1")

## ad the cluster
class.data.cluster <- class_data_baseline %>%
  left_join(bmi.cluster.mean, by = "Probanden_ID")

# exclude LEANER-15 and -24 because of missing data
class.data.cluster <- class.data.cluster %>%
  filter(Probanden_ID != "LEANER-15") %>%
  filter(Probanden_ID != "LEANER-25")
```

## Apply random forest to leaner data
```{r}
library(tidyverse)
library(caret)
library(randomForest)
df <- class.data.cluster
df$Probanden_ID <- NULL
df$Sex <- NULL
df$Visite <- NULL
df$BMI <- NULL
df$responder_mean <- as.factor(df$responder_mean) 

### save list for supplement table
library(writexl)
features <- colnames(df)
features_df <- data.frame(Feature = features)
#write_xlsx(features_df, "features_list.xlsx")

# Split the data into training and test set
set.seed(123)
training.samples <- df$responder_mean %>% 
  createDataPartition(p = 0.8, list = FALSE)
# In this case, 80% of the data will be used for training, and the remaining 20% will be used for the test set. 
train.data  <- df[training.samples, ]
test.data <- df[-training.samples, ]


# Define the control function for training with LOOCV and set random seeds for reproducebility
control <- trainControl(method = "LOOCV")

# Train the Random Forest model using LOOCV
model <- train(responder_mean ~ ., data = train.data,
               method = "rf", 
               trControl = control, 
               importance = TRUE)

# Print the model
print(model)

# Evaluate model performance
predicted_test <- predict(model, newdata = test.data[, -ncol(test.data)])

# Evaluate the model
confusionMatrix(predicted_test, as.factor(test.data$responder_mean)) -> confmat
confmat

# Feature importance plot
feature_importance <- importance(model$finalModel)
print(feature_importance)

###############
### plot feature importance
######################
feature_importance <- importance(model$finalModel)

# Create a data frame for easier visualization
feature_importance_df <- data.frame(
  feature = rownames(feature_importance),
  importance = feature_importance[, 4]  # Choose the correct importance metric
)

# Sort by importance
feature_importance_df <- feature_importance_df %>%
  arrange(desc(importance))

# Plot the feature importance

ggplot(feature_importance_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "steelblue") + 
  coord_flip() +  
  labs(title = "Feature importance in predicting cluster",
       x = "Feature",
       y = "Importance") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 24, color = "black")
  ) -> fig.3.feature.importance.model.long
fig.3.feature.importance.model.long

#ggsave("fig.3.feature.importance.model.long.pdf", plot = fig.3.feature.importance.model.long, width = 20, height = 20)

```

### Including BMI at baseline in the model
```{r}
df <- class.data.cluster
df$Probanden_ID <- NULL
df$Sex <- NULL
df$Visite <- NULL
#df$BMI <- NULL
df$responder_mean <- as.factor(df$responder_mean) 
# Split the data into training and test set
set.seed(123)
training.samples <- df$responder_mean %>% 
  createDataPartition(p = 0.8, list = FALSE)
# In this case, 80% of the data will be used for training, and the remaining 20% will be used for the test set. 
train.data  <- df[training.samples, ]
test.data <- df[-training.samples, ]


# Define the control function for training with LOOCV and set random seeds for reproducebility
control <- trainControl(method = "LOOCV")

# Train the Random Forest model using LOOCV
model_BMI <- train(responder_mean ~ ., data = train.data,
               method = "rf", 
               trControl = control, 
               importance = TRUE)

# Print the model
print(model_BMI)

# Evaluate model performance
predicted_test <- predict(model_BMI, newdata = test.data[, -ncol(test.data)])

# Evaluate the model
confusionMatrix(predicted_test, as.factor(test.data$responder_mean)) -> confmat_BMI
confmat_BMI

# Feature importance plot
feature_importance <- importance(model_BMI$finalModel)
print(feature_importance)

###############
### plot feature importance
######################
feature_importance <- importance(model_BMI$finalModel)

# Create a data frame for easier visualization
feature_importance_df <- data.frame(
  feature = rownames(feature_importance),
  importance = feature_importance[, 4]  # Choose the correct importance metric
)

# Sort by importance
feature_importance_df <- feature_importance_df %>%
  arrange(desc(importance))

# Plot the feature importance

ggplot(feature_importance_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "steelblue") +  # geom_col() is more direct for plotting pre-summarized values
  coord_flip() +  # Horizontal bar plot for better label readability
  labs(title = "Feature importance in predicting cluster",
       x = "Feature",
       y = "Importance") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 24, color = "black")
  ) -> fig.3.feature.importance.model.long.bmi
fig.3.feature.importance.model.long.bmi

ggsave("fig.3.feature.importance.model.long.bmi.pdf", plot = fig.3.feature.importance.model.long.bmi, width = 20, height = 20)
```



### Set up short model
```{r}
# # Define the control function for training with LOOCV and set random seeds for reproducebility
seeds.used <- readRDS("rowdata/classifier/seeds.rds")
control <- trainControl(method = "LOOCV",
                        seeds = seeds.used)

model.short <- train(responder_mean ~ Oscillibacter.sp..57_20. + 
                       NA.Faecalibacterium.species.incertae.sedis + 
                       ldl + 
                       sys_mean, 
                     data = train.data,
               method = "rf", 
               trControl = control, 
               importance = TRUE)

# Predict on test data, excluding the last column with the disease status
predicted_test <- predict(model.short, newdata = test.data[, -ncol(test.data)])

# Evaluate the model
confusionMatrix(predicted_test, as.factor(test.data$responder_mean)) -> confmat.short
confmat.short

########
## plot confusion matrix
#############
# 1) Define the confusion matrix
conf_mat <- matrix(
  c(4, 1,
    0, 1),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    c("Predicted Non-Responder", "Predicted Responder"),  # rows
    c("Actual Non-Responder",    "Actual Responder")      # columns
  )
)

# 2) Convert to a data frame for ggplot
conf_mat_df <- as.data.frame.table(conf_mat)
colnames(conf_mat_df) <- c("Prediction", "Reference", "Freq")

# 3) Mark each cell as Correct vs. Incorrect (diagonal vs. off-diagonal)
conf_mat_df$Color <- ifelse(
  (conf_mat_df$Prediction == "Predicted Non-Responder" & 
   conf_mat_df$Reference  == "Actual Non-Responder") |
  (conf_mat_df$Prediction == "Predicted Responder" & 
   conf_mat_df$Reference  == "Actual Responder"),
  "Correct",
  ifelse(conf_mat_df$Freq == 0, "No Data", "Incorrect")
)

# 4) Plot heatmap with ggplot
ggplot(conf_mat_df, aes(x = Reference, y = Prediction, fill = Color)) +
  geom_tile(color = "black") +
  geom_text(aes(label = Freq), color = "black", size = 8) +
  scale_fill_manual(
    "Classification",
    values = c("Correct"   = "#5ab4ac",  # teal
               "Incorrect" = "#e9a3c9",  # pink
               "No Data"   = "white")    # zero-frequency cells
  ) +
  coord_fixed() +
  labs(x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(
    panel.border    = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "top",
    legend.title    = element_text(size = 24),
    legend.text     = element_text(size = 24),
    axis.text       = element_text(size = 24, color = "black"),
    axis.title      = element_text(size = 24, face = "bold"),
    plot.title      = element_text(size = 24, face = "bold", hjust = 0.5)
  ) -> fig.3.confusion.matrix.leaner

#ggsave("fig.3.confusion.matrix.leaner.pdf", plot = fig.3.confusion.matrix.leaner, width = 12, height = 10)


#######################
##### feature importance
########################


 #Access the underlying random forest model
rf_model <- model.short$finalModel

# Extract the Gini importance (Mean Decrease Gini)
gini_importance <- importance(rf_model, type = 2)  # Type 2 is Mean Decrease Gini

# Convert to data frame for ggplot
gini_importance_df <- data.frame(
  Feature = rownames(gini_importance),
  Importance = gini_importance[, "MeanDecreaseGini"]
)

# Plot Gini importance
gini <- gini_importance_df
gini$Feature  <- gsub(pattern = "\\.", replace = " ", x = gini$Feature )
gini$Feature  <- gsub(pattern = "^NA", replace = "", x = gini$Feature )
gini$Feature <- gsub(pattern = "species", replace = "sp", x = gini$Feature )
gini$Feature <- gsub(pattern = "sys_mean", replace = "sys BP", x = gini$Feature )
gini$Feature <- gsub(pattern = "sp", replace = "sp.", x = gini$Feature )

ggplot(gini, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +  # Flip coordinates for better readability
  theme_minimal() +
  labs(
    x = "Features",
    y = "Feature importance gini index"
  ) +
  ylim(c(0, max(gini$Importance) + 1)) +  # Set y-limits to be a bit higher than max value
  theme(
    axis.text = element_text(size = 24, color = "black"),
    axis.title = element_text(size = 24, color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),  # Add border around the plot
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    legend.position = "none"  # Remove legend
  ) +
  scale_fill_manual(values = c("#fc8d59", "#0072B2", "#FFA8BB", "#00998F"))  -> fig.3.feature.importance.model.short
fig.3.feature.importance.model.short
```

### short model with BMI
```{r}
seeds.used <- readRDS("rowdata/classifier/seeds.rds")
control <- trainControl(method = "LOOCV",
                        seeds = seeds.used)

model.short.bmi <- train(responder_mean ~ BMI + 
                           Oscillibacter.sp..57_20. + 
                       NA.Faecalibacterium.species.incertae.sedis + 
                       ldl + 
                       sys_mean, 
                     data = train.data,
               method = "rf", 
               trControl = control, 
               importance = TRUE)

predicted_test <- predict(model.short.bmi, newdata = test.data[, -ncol(test.data)])

# Evaluate the model
confusionMatrix(predicted_test, as.factor(test.data$responder_mean)) -> confmat.short.bmi
confmat.short.bmi
#ggsave("confmat.short.bmi.pdf", plot = confmat.short.bmi, width = 12, height = 10)

########
## plot confusion matrix
#############
# 1) Define the confusion matrix
conf_mat <- matrix(
  c(4, 1,
    0, 1),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    c("Predicted Non-Responder", "Predicted Responder"),  # rows
    c("Actual Non-Responder",    "Actual Responder")      # columns
  )
)

# 2) Convert to a data frame for ggplot
conf_mat_df <- as.data.frame.table(conf_mat)
colnames(conf_mat_df) <- c("Prediction", "Reference", "Freq")

# 3) Mark each cell as Correct vs. Incorrect (diagonal vs. off-diagonal)
conf_mat_df$Color <- ifelse(
  (conf_mat_df$Prediction == "Predicted Non-Responder" & 
   conf_mat_df$Reference  == "Actual Non-Responder") |
  (conf_mat_df$Prediction == "Predicted Responder" & 
   conf_mat_df$Reference  == "Actual Responder"),
  "Correct",
  ifelse(conf_mat_df$Freq == 0, "No Data", "Incorrect")
)

# 4) Plot heatmap with ggplot
ggplot(conf_mat_df, aes(x = Reference, y = Prediction, fill = Color)) +
  geom_tile(color = "black") +
  geom_text(aes(label = Freq), color = "black", size = 8) +
  scale_fill_manual(
    "Classification",
    values = c("Correct"   = "#5ab4ac",  # teal
               "Incorrect" = "#e9a3c9",  # pink
               "No Data"   = "white")    # zero-frequency cells
  ) +
  coord_fixed() +
  labs(x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(
    panel.border    = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "top",
    legend.title    = element_text(size = 16),
    legend.text     = element_text(size = 16),
    axis.text       = element_text(size = 16, color = "black"),
    axis.title      = element_text(size = 16, face = "bold"),
    plot.title      = element_text(size = 16, face = "bold", hjust = 0.5)
  ) -> fig.3.confusion.matrix.leaner.bmi
fig.3.confusion.matrix.leaner.bmi

#ggsave("fig.3.confusion.matrix.leaner.bmi.pdf", plot = fig.3.confusion.matrix.leaner.bmi, width = 12, height = 10)

#######################
##### feature importance
########################

rf_model <- model.short.bmi$finalModel

gini_importance <- importance(rf_model, type = 2)  # Type 2 is Mean Decrease Gini

gini_importance_df <- data.frame(
  Feature = rownames(gini_importance),
  Importance = gini_importance[, "MeanDecreaseGini"]
)

gini <- gini_importance_df
gini$Feature  <- gsub(pattern = "\\.", replace = " ", x = gini$Feature )
gini$Feature  <- gsub(pattern = "^NA", replace = "", x = gini$Feature )

gini <- gini %>%
  mutate(
    Feature = str_trim(Feature),  # remove leading/trailing spaces
    Feature = recode(Feature,
      "Oscillibacter sp  57_20" = "Oscillibacter sp. 57_20",
      "Faecalibacterium species incertae sedis" = "Faecalibacterium sp. incertae sedis",
      "ldl" = "LDL",
      "sys_mean" = "Sys BP"
    )
  )

feature_labels <- c(
  "Oscillibacter sp. 57_20" = "Oscilli sp.\n57_20",
  "Faecalibacterium sp. incertae sedis" = "Faecali sp.\nincertae sedis",
  "Sys BP" = "Sys BP",
  "LDL" = "LDL"
)

ggplot(gini, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +
  scale_x_discrete(labels = feature_labels) +  # custom y-axis with line breaks
  theme_minimal() +
  labs(
    x = "Features",
    y = "Feature importance (Gini index)"
  ) +
  ylim(c(0, max(gini$Importance) + 1)) +
  theme(
    axis.text = element_text(size = 24, color = "black"),
    axis.title = element_text(size = 24, color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  ) +
  scale_fill_manual(values = c("#fc8d59", "grey", "#0072B2", "#FFA8BB", "#00998F")) -> fig.3.feature.importance.model.short.bmi
ggsave("fig.3.feature.importance.model.short.bmi.pdf", plot = fig.3.feature.importance.model.short.bmi, width = 8, height = 8)
fig.3.feature.importance.model.short.bmi
```


### Calculating ROC curve
```{r}
# ROC curve: what are the probabilities for each classification?
# what are the probabilities?
predicted_probs_leaner <- predict(model.short, newdata = test.data, type = "prob")

# Check the structure of predicted probabilities
str(predicted_probs_leaner)

samp.leaner <- class.data.cluster %>%
  select(Probanden_ID, responder_mean)

samp.leaner <- samp.leaner[rownames(samp.leaner) %in% rownames(predicted_probs_leaner), ]

combined_data_leaner <- cbind(samp.leaner, predicted_probs_leaner)
colnames(combined_data_leaner) <- c("sample", "responder_calculated", "0", "1")

# Check the combined data
head(combined_data_leaner)

# Create a prediction object
library(ROCR)
pred_leaner <- prediction(combined_data_leaner$`1`, combined_data_leaner$responder_calculated)

# Compute the performance measures
perf <- performance(pred_leaner, measure = "tpr", x.measure = "fpr")

# Extract the data for plotting
roc_data_leaner <- data.frame(
  FPR = perf@x.values[[1]],
  TPR = perf@y.values[[1]],
  Cutoff = perf@alpha.values[[1]]
)

# Calculate and print the AUC
auc <- performance(pred_leaner, measure = "auc")
auc <- auc@y.values[[1]]
print(paste("Area Under the ROC Curve:", round(auc, 2)))

# Plot the ROC curve
ggplot(roc_data_leaner, aes(x = FPR, y = TPR)) +
  geom_line(size = 1, color = "steelblue") +
  geom_abline(linetype = "dashed", color = "red") +  # Add diagonal line
  annotate("text", x = 0.6, y = 0.2, label = paste("AUC =", round(auc, 2)), size = 10) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), name = "False Positive Rate") +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), name = "True Positive Rate") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 24, color = "black"),
    axis.text = element_text(size = 24, color = "black"),
    legend.position = "bottom",
    panel.border = element_rect(color = "black", fill = NA,  size = 1),
    axis.line = element_blank()
  ) -> fig.3.ROC.LEANER

```

## Repeated 80/20 Monte-Carlo CV (100 repeats)
repeat the 80/20 split many times; train each time (with inner CV for tuning), predict on the held-out 20%, collect metrics and feature importances; then summarize.

```{r}
## ===============================
## Random Forest: Repeated CV + Nested CV 
## ===============================
library(caret)
library(pROC)
library(randomForest)


#####################
####### Repeated 80/20 with inner CV
#############################
# Outer split: random 80% train / 20% test, repeated 100 times.
# Inner CV: 5-fold inside each 80% to pick the best mtry.
# What you get: a distribution of performance (median, IQR), showing stability/variance across many random splits.
# → Answers “Could our result be a lucky split?”


set.seed(42)

## ---- 0) DATA PREP 
dat <- df %>%
  select(!BMI) %>%
  mutate(
    responder_mean = factor(
      ifelse(as.character(responder_mean) %in% c("1", 1, "Responder"),
             "Responder", "NonResponder"),
      levels = c("Responder", "NonResponder")
    )
  ) %>%
  drop_na()

# number of predictors
p <- ncol(dat) - 1

##############################################
## ---- 1) robust varImp extractor (builds an Overall column)
#####################################
### after training the model, this function pulls variable importance from caret::varImp() and returns a table with feature, its importance and how often it came up 

get_vi <- function(fit, iter_or_fold){
  vi_obj <- caret::varImp(fit, scale = FALSE)
  imp_raw <- tryCatch(vi_obj$importance, error = function(e) as.data.frame(vi_obj))
  imp_df <- as.data.frame(imp_raw)

  if(!"Overall" %in% colnames(imp_df)){
    if(ncol(imp_df) > 1){
      imp_df$Overall <- rowMeans(imp_df, na.rm = TRUE)
    } else {
      colnames(imp_df) <- "Overall"
    }
  }

  tibble(
    Feature    = rownames(imp_df),
    Importance = imp_df[ , "Overall", drop = TRUE],
    iter       = iter_or_fold
  )
}


#######################################
## ---- 2) Common training controls & grid ----
##########################################
# mtry is the RF hyperparameter (how many features are considered at each split)
# inner 5-fold CV = we try different mtry values and pick the one with the best ROC

rf_ntree <- 1000
rf_grid  <- expand.grid(mtry = seq(1, min(p, floor(sqrt(p)) + 2), by = 1))

ctrl_inner <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,  
  savePredictions = "final",
  allowParallel = TRUE
)

##########################
## A) Repeated 80/20 Monte-Carlo CV (FULL MODEL)
#########################
# we do the resampling for 100 times

B <- 100
metrics_mc  <- vector("list", B)
imp_mc_list <- vector("list", B)

for(b in seq_len(B)){
  # stratified 80/20 split
  tr_idx <- createDataPartition(dat$responder_mean, p = 0.8, list = FALSE)
  tr <- dat[tr_idx, ]
  te <- dat[-tr_idx, ]

  # train on 80% with inner CV (optimize ROC)
  fit_b <- train(
    responder_mean ~ .,
    data = tr,
    method = "rf",
    metric = "ROC",
    trControl = ctrl_inner,
    tuneGrid = rf_grid,
    ntree = rf_ntree,
    importance = TRUE
  )

  # predict probs on held-out 20%
  prob <- predict(fit_b, newdata = te, type = "prob")[, "Responder"]
  pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                 levels = c("Responder", "NonResponder"))

  # metrics (positive class = Responder)
  roc_obj <- pROC::roc(response = te$responder_mean,
                       predictor = prob,
                       levels = c("NonResponder","Responder"),
                       direction = "<")
  cm <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")

  metrics_mc[[b]] <- tibble(
    iter      = b,
    AUC       = as.numeric(pROC::auc(roc_obj)),
    Accuracy  = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    Best_mtry = fit_b$bestTune$mtry
  )

  imp_mc_list[[b]] <- get_vi(fit_b, b)
}

### summary of the result, save AUC. Accuaracy, Sensitivity, Specificity and which mtry was chosen
mc_metrics     <- bind_rows(metrics_mc)
mc_importance  <- bind_rows(imp_mc_list)

## Summaries you can report (median/IQR over repeats)
## gives median and IQR of the key metrics across the 100 repeats → this shows typical performance and its spread.
mc_summary <- mc_metrics %>%
  summarise(
    AUC_median = median(AUC, na.rm=TRUE),
    AUC_IQR_L  = quantile(AUC, 0.25, na.rm=TRUE),
    AUC_IQR_U  = quantile(AUC, 0.75, na.rm=TRUE),
    Acc_median = median(Accuracy, na.rm=TRUE),
    Sens_median= median(Sensitivity, na.rm=TRUE),
    Spec_median= median(Specificity, na.rm=TRUE)
  )
print(mc_summary)

## Feature stability: Top-k frequency across repeats
# median and IQR of the key metrics across the 100 repeats → this shows typical performance and its spread.
k <- 5
mc_stability <- mc_importance %>%
  group_by(iter) %>%
  slice_max(order_by = Importance, n = k, with_ties = FALSE) %>%
  ungroup() %>%
  count(Feature, name = "TopK_hits") %>%
  arrange(desc(TopK_hits)) %>%
  mutate(TopK_freq = TopK_hits / B)
mc_stability %>% print(n = 30)

mc_mean_importance <- mc_importance %>%
  group_by(Feature) %>%
  summarise(MeanImportance = mean(Importance, na.rm=TRUE),
            SDImportance   = sd(Importance, na.rm=TRUE)) %>%
  arrange(desc(MeanImportance))
mc_mean_importance 

## AUC distribution plot for supplement
ggplot(mc_metrics, aes(x = AUC)) + geom_histogram(bins = 20) + theme_minimal()

#(1) cleaned labels, (2) defined a tuning grid, (3) repeated a stratified 80/20 split 100 times, (4) tuned mtry inside the 80% via 5-fold CV each time, (5) tested on the 20%, (6) saved metrics and feature importances, and (7) summarized medians/IQRs and feature stability.
```

```{r}
## ===============================
## B) Nested CV (FULL MODEL) — outer 5-fold, inner 5-fold
## ===============================
# Outer split: 5 folds (≈20% test each), cycling so every sample is test once.
# Inner CV: the same 5-fold tuning inside each outer training fold.
# what we get: one unbiased estimate (mean ± SD across the 5 outer folds).
# → Answers “What’s our overall generalization when every point serves as test exactly once?”
# this should give the same results then the repeated 80/20 with inner CV. We will keep repeated 80/20 for the distribution, and mention that we confirmed with a 5x5 nested CV giving similar mean +- SD


K_outer <- 5
outer_folds <- createFolds(dat$responder_mean, k = K_outer, list = TRUE, returnTrain = TRUE)

metrics_nested <- vector("list", K_outer)
imp_nested    <- vector("list", K_outer)

for(k in seq_len(K_outer)){
  tr <- dat[ outer_folds[[k]], ]
  te <- dat[-outer_folds[[k]], ]

  fit_k <- train(
    responder_mean ~ .,
    data = tr,
    method = "rf",
    metric = "ROC",
    trControl = ctrl_inner,   # inner CV happens here
    tuneGrid = rf_grid,
    ntree = rf_ntree,
    importance = TRUE
  )

  prob <- predict(fit_k, newdata = te, type = "prob")[, "Responder"]
  pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                 levels = c("Responder", "NonResponder"))

  roc_k <- pROC::roc(response = te$responder_mean,
                     predictor = prob,
                     levels = c("NonResponder","Responder"),
                     direction = "<")
  cm_k  <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")

  metrics_nested[[k]] <- tibble(
    fold      = k,
    AUC       = as.numeric(pROC::auc(roc_k)),
    Accuracy  = as.numeric(cm_k$overall["Accuracy"]),
    Sensitivity = as.numeric(cm_k$byClass["Sensitivity"]),
    Specificity = as.numeric(cm_k$byClass["Specificity"]),
    Best_mtry = fit_k$bestTune$mtry
  )

  imp_nested[[k]] <- get_vi(fit_k, k)
}

nested_perf <- bind_rows(metrics_nested)
nested_vi   <- bind_rows(imp_nested)

## Unbiased performance estimate (mean ± SD across outer folds)
nested_summary <- nested_perf %>%
  summarise(
    AUC_mean = mean(AUC), AUC_sd = sd(AUC),
    Acc_mean = mean(Accuracy), Acc_sd = sd(Accuracy),
    Sens_mean= mean(Sensitivity), Sens_sd = sd(Sensitivity),
    Spec_mean= mean(Specificity), Spec_sd = sd(Specificity)
  )
print(nested_summary)

## Feature stability across outer folds (Top-5 frequency)
nested_stability <- nested_vi %>%
  group_by(iter) %>%  # here iter == outer fold id
  mutate(rank = rank(-Importance, ties.method = "first")) %>%
  ungroup() %>%
  count(Feature, wt = (rank <= 5), name = "Top5_hits") %>%
  mutate(Top5_freq = Top5_hits / K_outer) %>%
  arrange(desc(Top5_freq), desc(Top5_hits))
nested_stability %>% print(n = 30)

```


```{r}
## ===============================
## SHORT MODEL variants
## ===============================
## Use your 4 features 

dat <- df %>%
  select(!BMI) %>%
  mutate(
    responder_mean = factor(
      ifelse(as.character(responder_mean) %in% c("1", 1, "Responder"),
             "Responder", "NonResponder"),
      levels = c("Responder", "NonResponder")
    )
  ) %>%
  drop_na()

short_features <- c("Oscillibacter.sp..57_20.",
                     "NA.Faecalibacterium.species.incertae.sedis", "ldl",
                     "sys_mean")

  dat_short <- dat %>% select(all_of(c(short_features, "responder_mean")))

  #(Re)use rf_grid tailored to number of predictors
  p_short <- ncol(dat_short) - 1
  rf_grid_short <- expand.grid(mtry = seq(1, min(p_short, floor(sqrt(p_short)) + 2), by = 1))

 # ---- Repeated MC CV for short model ----
  B <- 100
  metrics_mc_short  <- vector("list", B)
  imp_mc_short_list <- vector("list", B)
  for(b in seq_len(B)){
    tr_idx <- createDataPartition(dat_short$responder_mean, p = 0.8, list = FALSE)
    tr <- dat_short[tr_idx, ]; te <- dat_short[-tr_idx, ]
    fit_b <- train(responder_mean ~ ., data = tr, method = "rf", metric = "ROC",
                   trControl = ctrl_inner, tuneGrid = rf_grid_short,
                   ntree = rf_ntree, importance = TRUE)
    prob <- predict(fit_b, newdata = te, type = "prob")[, "Responder"]
    pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                   levels = c("Responder","NonResponder"))
    roc_obj <- pROC::roc(te$responder_mean, prob, levels = c("NonResponder","Responder"), direction = "<")
    cm <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")
    metrics_mc_short[[b]] <- tibble(iter=b, AUC=as.numeric(pROC::auc(roc_obj)),
                                    Accuracy=as.numeric(cm$overall["Accuracy"]),
                                    Sensitivity=as.numeric(cm$byClass["Sensitivity"]),
                                    Specificity=as.numeric(cm$byClass["Specificity"]),
                                    Best_mtry = fit_b$bestTune$mtry)
    imp_mc_short_list[[b]] <- get_vi(fit_b, b)
  }
  mc_metrics_short    <- bind_rows(metrics_mc_short)
  mc_importance_short <- bind_rows(imp_mc_short_list)
```


### train model on full data for application in another dataset
```{r}
# Train final model on full data
final_fit <- train(
  responder_mean ~ ., 
  data = dat_short, 
  method = "rf", 
  metric = "ROC",
  trControl = ctrl_inner, 
  tuneGrid = rf_grid_short,
  ntree = rf_ntree,
  importance = TRUE
)


###### train a model without ldl
dat_short_no_ldl <- dat_short %>%
  select(!ldl)

final_fit_no_ldl <- train(
  responder_mean ~ ., 
  data = dat_short_no_ldl, 
  method = "rf", 
  metric = "ROC",
  trControl = ctrl_inner, 
  tuneGrid = rf_grid_short,
  ntree = rf_ntree,
  importance = TRUE
)

```

## Repeated 80/20 CV - performance summary + ploots + stability

```{r}
# Median [IQR] across 100 repeats
mc_summary_short <- mc_metrics_short %>%
  summarise(
    AUC_median = median(AUC, na.rm=TRUE),
    AUC_IQR_L  = quantile(AUC, 0.25, na.rm=TRUE),
    AUC_IQR_U  = quantile(AUC, 0.75, na.rm=TRUE),
    Acc_median = median(Accuracy, na.rm=TRUE),
    Acc_IQR_L  = quantile(Accuracy, 0.25, na.rm=TRUE),
    Acc_IQR_U  = quantile(Accuracy, 0.75, na.rm=TRUE),
    Sens_median= median(Sensitivity, na.rm=TRUE),
    Sens_IQR_L = quantile(Sensitivity, 0.25, na.rm=TRUE),
    Sens_IQR_U = quantile(Sensitivity, 0.75, na.rm=TRUE),
    Spec_median= median(Specificity, na.rm=TRUE),
    Spec_IQR_L = quantile(Specificity, 0.25, na.rm=TRUE),
    Spec_IQR_U = quantile(Specificity, 0.75, na.rm=TRUE)
  )
print(mc_summary_short)

# 95% percentile CIs (optional, for supplement)
mc_ci_short <- mc_metrics_short %>%
  summarise(
    AUC_CI_L  = quantile(AUC, 0.025, na.rm=TRUE),
    AUC_CI_U  = quantile(AUC, 0.975, na.rm=TRUE),
    Acc_CI_L  = quantile(Accuracy, 0.025, na.rm=TRUE),
    Acc_CI_U  = quantile(Accuracy, 0.975, na.rm=TRUE)
  )
print(mc_ci_short)

# Distribution of mtry chosen in inner CV
mtry_dist_short <- mc_metrics_short %>%
  count(Best_mtry, name = "n") %>%
  mutate(freq = n / sum(n)) %>%
  arrange(desc(n))
print(mtry_dist_short)

# Top-k feature stability over repeats (k=all features here = 5)
k <- 5
mc_stability_short <- mc_importance_short %>%
  group_by(iter) %>%
  slice_max(order_by = Importance, n = k, with_ties = FALSE) %>%
  ungroup() %>%
  count(Feature, name = "TopK_hits") %>%
  mutate(TopK_freq = TopK_hits / n_distinct(mc_importance_short$iter)) %>%
  arrange(desc(TopK_freq), desc(TopK_hits))
print(mc_stability_short)

# Mean importance across repeats (helps ranking)
mc_mean_importance_short <- mc_importance_short %>%
  group_by(Feature) %>%
  summarise(MeanImportance = mean(Importance, na.rm=TRUE),
            SDImportance   = sd(Importance, na.rm=TRUE)) %>%
  arrange(desc(MeanImportance))
print(mc_mean_importance_short)

# Quick plots (optional)
ggplot(mc_metrics_short, aes(x = AUC)) +
  geom_histogram(bins = 20) + theme_minimal() +
  labs(title = "Short model: AUC distribution (100 repeats)") -> AUC.distribution.short
ggsave(AUC.distribution.short, file = "AUC.distribution.short.pdf", width = 5, height = 5)


ggplot(mc_metrics_short, aes(y = AUC, x = 1)) +
  geom_boxplot(width = 0.2) + theme_minimal() +
  labs(title = "Short model: AUC (median[IQR])", x = NULL) -> Median.AUC.short
ggsave(Median.AUC.short, file = "Median.AUC.short.pdf", width = 5, height = 5)


# prevalence and class balance
table(dat_short$responder_mean)
prop.table(table(dat_short$responder_mean))



## balances accurracy
# Re-run repeated 80/20 for short model to also collect predictions
B <- 100
metrics_mc_short  <- vector("list", B)
imp_mc_short_list <- vector("list", B)
pred_store_short  <- vector("list", B)

for(b in seq_len(B)){
  tr_idx <- createDataPartition(dat_short$responder_mean, p = 0.8, list = FALSE)
  tr <- dat_short[tr_idx, ]; te <- dat_short[-tr_idx, ]

  fit_b <- train(responder_mean ~ ., data = tr, method = "rf", metric = "ROC",
                 trControl = ctrl_inner, tuneGrid = rf_grid_short,
                 ntree = rf_ntree, importance = TRUE)

  prob <- predict(fit_b, newdata = te, type = "prob")[, "Responder"]
  pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                 levels = c("Responder","NonResponder"))

  roc_obj <- pROC::roc(te$responder_mean, prob,
                       levels = c("NonResponder","Responder"), direction = "<")
  cm <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")

  # confusion-matrix-derived metrics
  TN <- cm$table["NonResponder","NonResponder"]; TP <- cm$table["Responder","Responder"]
  FP <- cm$table["Responder","NonResponder"];    FN <- cm$table["NonResponder","Responder"]
  Precision <- ifelse((TP+FP)==0, NA, TP/(TP+FP))
  Recall    <- ifelse((TP+FN)==0, NA, TP/(TP+FN))  # = Sensitivity
  F1        <- ifelse(is.na(Precision) | is.na(Recall) | (Precision+Recall)==0, NA,
                      2*Precision*Recall/(Precision+Recall))
  BalAcc    <- (as.numeric(cm$byClass["Sensitivity"]) + as.numeric(cm$byClass["Specificity"]))/2
  MCC       <- ifelse((TP+FP)==0 | (TP+FN)==0 | (TN+FP)==0 | (TN+FN)==0, NA,
                      ((TP*TN - FP*FN) / sqrt((TP+FP)*(TP+FN)*(TN+FP)*(TN+FN))))

  metrics_mc_short[[b]] <- tibble(
    iter=b,
    AUC=as.numeric(pROC::auc(roc_obj)),
    Accuracy=as.numeric(cm$overall["Accuracy"]),
    Sensitivity=as.numeric(cm$byClass["Sensitivity"]),
    Specificity=as.numeric(cm$byClass["Specificity"]),
    Precision=Precision, Recall=Recall, F1=F1,
    BalancedAccuracy = BalAcc,
    MCC = MCC,
    Best_mtry = fit_b$bestTune$mtry
  )

  imp_mc_short_list[[b]] <- get_vi(fit_b, b)
  pred_store_short[[b]]  <- tibble(iter=b, truth=te$responder_mean, prob=prob, pred=pred)
}

mc_metrics_short    <- bind_rows(metrics_mc_short)
mc_importance_short <- bind_rows(imp_mc_short_list)
preds_short         <- bind_rows(pred_store_short)

# Summaries to report
mc_summary_short_ext <- mc_metrics_short %>%
  summarise(
    AUC_median = median(AUC, na.rm=TRUE),
    AUC_IQR_L  = quantile(AUC, 0.25, na.rm=TRUE),
    AUC_IQR_U  = quantile(AUC, 0.75, na.rm=TRUE),
    Acc_median = median(Accuracy, na.rm=TRUE),
    BalAcc_median = median(BalancedAccuracy, na.rm=TRUE),
    F1_median  = median(F1, na.rm=TRUE),
    MCC_median = median(MCC, na.rm=TRUE)
  )
print(mc_summary_short_ext)


### calculate specicifiy and accuarcy
mc_summary_short_meansd <- mc_metrics_short %>%
  summarise(
    AUC_mean = mean(AUC, na.rm=TRUE),       AUC_sd = sd(AUC, na.rm=TRUE),
    Acc_mean = mean(Accuracy, na.rm=TRUE),  Acc_sd = sd(Accuracy, na.rm=TRUE),
    Sens_mean= mean(Sensitivity, na.rm=TRUE), Sens_sd = sd(Sensitivity, na.rm=TRUE),
    Spec_mean= mean(Specificity, na.rm=TRUE), Spec_sd = sd(Specificity, na.rm=TRUE)
  )

# Format into "mean ± sd" strings
perf_table_repeated <- tibble::tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity"),
  Value  = c(
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$AUC_mean, mc_summary_short_meansd$AUC_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Acc_mean, mc_summary_short_meansd$Acc_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Sens_mean, mc_summary_short_meansd$Sens_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Spec_mean, mc_summary_short_meansd$Spec_sd)
  )
)


### precision-recall AUC 
###thats a good way to adress the issue that we have an unbalanced dataset (1/3 responder. 2/3 non repsonder). AP is stricter that ROC-AUC: it asks when the model calls someone a responder, how often is it correct, and how completly does it find them?

library(PRROC)
ap_tbl <- preds_short %>%
  group_by(iter) %>%
  summarise(
    AP = {
      pos <- prob[truth=="Responder"]; neg <- prob[truth=="NonResponder"]
      if(length(pos)>0 && length(neg)>0)
        PRROC::pr.curve(scores.class0 = pos, scores.class1 = neg, curve = FALSE)$auc.integral
      else NA_real_
    },
    .groups="drop"
  )
ap_summary <- ap_tbl %>%
  summarise(AP_median = median(AP, na.rm=TRUE),
            AP_IQR_L  = quantile(AP, 0.25, na.rm=TRUE),
            AP_IQR_U  = quantile(AP, 0.75, na.rm=TRUE))
print(ap_summary)

# The short model achieved a median average precision of 0.80 [0.67–1.00], substantially higher than the responder prevalence (~0.38). This indicates that, when the model predicts a responder, it is correct most of the time, although performance varied across splits.


## rank satbility
rankstab_short <- mc_importance_short %>%
  group_by(iter) %>%
  arrange(desc(Importance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  group_by(Feature) %>%
  summarise(
    MedianRank = median(rank),
    IQR_L = quantile(rank, 0.25),
    IQR_U = quantile(rank, 0.75),
    .groups="drop"
  ) %>%
  arrange(MedianRank)
print(rankstab_short)

#rankstab_short
# Across 100 resamples, feature ranking was stable: Oscillibacter was consistently the top predictor (median rank 1 [IQR 1–2]), followed by Faecalibacterium (2 [1–2]). Systolic BP ranked stably third, while LDL was always last. This reinforces that prediction in the short model was mainly driven by the two microbiome features.

```

```{r}
## ===== Short model performance: mean ± SD =====
mc_summary_short_meansd <- mc_metrics_short %>%
  summarise(
    AUC_mean = mean(AUC, na.rm=TRUE),       AUC_sd = sd(AUC, na.rm=TRUE),
    Acc_mean = mean(Accuracy, na.rm=TRUE),  Acc_sd = sd(Accuracy, na.rm=TRUE),
    Sens_mean= mean(Sensitivity, na.rm=TRUE), Sens_sd = sd(Sensitivity, na.rm=TRUE),
    Spec_mean= mean(Specificity, na.rm=TRUE), Spec_sd = sd(Specificity, na.rm=TRUE),
    BalAcc_mean = mean(BalancedAccuracy, na.rm=TRUE), BalAcc_sd = sd(BalancedAccuracy, na.rm=TRUE),
    F1_mean = mean(F1, na.rm=TRUE), F1_sd = sd(F1, na.rm=TRUE),
    MCC_mean = mean(MCC, na.rm=TRUE), MCC_sd = sd(MCC, na.rm=TRUE)
  )

perf_short <- tibble::tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity","Balanced Accuracy","F1","MCC"),
  Value  = c(
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$AUC_mean, mc_summary_short_meansd$AUC_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Acc_mean, mc_summary_short_meansd$Acc_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Sens_mean, mc_summary_short_meansd$Sens_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$Spec_mean, mc_summary_short_meansd$Spec_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$BalAcc_mean, mc_summary_short_meansd$BalAcc_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$F1_mean, mc_summary_short_meansd$F1_sd),
    sprintf("%.2f ± %.2f", mc_summary_short_meansd$MCC_mean, mc_summary_short_meansd$MCC_sd)
  )
)

print(perf_short)


## ===== Short model feature importance + rank stability =====
feature_table_short <- mc_importance_short %>%
  group_by(Feature) %>%
  summarise(
    MeanImportance = mean(Importance, na.rm=TRUE),
    SDImportance   = sd(Importance, na.rm=TRUE),
    .groups="drop"
  )

rankstab_short <- mc_importance_short %>%
  group_by(iter) %>%
  arrange(desc(Importance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  group_by(Feature) %>%
  summarise(
    MedianRank = median(rank),
    IQR_L = quantile(rank, 0.25),
    IQR_U = quantile(rank, 0.75),
    .groups="drop"
  )

features_summary_short <- left_join(feature_table_short, rankstab_short, by = "Feature") %>%
  mutate(
    `Importance (mean ± SD)` = sprintf("%.2f ± %.2f", MeanImportance, SDImportance),
    `Rank (median [IQR])`    = sprintf("%.0f [%.0f–%.0f]", MedianRank, IQR_L, IQR_U)
  ) %>%
  arrange(MedianRank) %>%                         # <-- arrange while MedianRank still exists
  select(Feature, `Importance (mean ± SD)`, `Rank (median [IQR])`)

print(features_summary_short)

```

## nested CV to validate results from the short model
```{r}
## ===============================
## Nested CV for SHORT MODEL (caret)
## Outer 5-fold, Inner 5-fold (or 3-fold)
## ===============================

library(tidyverse)
library(caret)
library(pROC)
library(randomForest)

set.seed(2025)

## ---- 0) Build the short-model dataset (4 features + outcome) ----
short_features <- c("Oscillibacter.sp..57_20.",
                    "NA.Faecalibacterium.species.incertae.sedis",
                    "ldl", "sys_mean")

dat_short <- df %>%
  # ensure valid two-class factor with positive class first
  mutate(
    responder_mean = factor(
      ifelse(as.character(responder_mean) %in% c("1", 1, "Responder"),
             "Responder", "NonResponder"),
      levels = c("Responder", "NonResponder")
    )
  ) %>%
  select(all_of(c(short_features, "responder_mean"))) %>%
  drop_na()

## ---- 1) Random Forest settings ----
rf_ntree <- 1000
p_short  <- ncol(dat_short) - 1
rf_grid_short <- expand.grid(mtry = seq(1, min(p_short, floor(sqrt(p_short)) + 2), by = 1))
# With 4 features this is typically mtry = 1:4

## ---- 2) Inner CV control (choose ONE) ----
# A) Inner 5-fold (default)
ctrl_inner5 <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,  # optimize ROC
  savePredictions = "final",
  allowParallel = TRUE
)

# B) Inner 3-fold (recommended if n is very small)
ctrl_inner3 <- trainControl(
  method = "cv", number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

## ---- 3) Helper: robust varImp extractor ----
get_vi <- function(fit, outer_id){
  vi_obj <- caret::varImp(fit, scale = FALSE)
  imp_raw <- tryCatch(vi_obj$importance, error = function(e) as.data.frame(vi_obj))
  imp_df  <- as.data.frame(imp_raw)
  if(!"Overall" %in% colnames(imp_df)){
    if(ncol(imp_df) > 1) imp_df$Overall <- rowMeans(imp_df, na.rm = TRUE) else colnames(imp_df) <- "Overall"
  }
  tibble(
    Feature    = rownames(imp_df),
    Importance = imp_df[, "Overall", drop = TRUE],
    fold       = outer_id
  )
}

## ---- 4) Outer 5-fold splits (stratified) ----
K_outer <- 5
outer_folds <- createFolds(dat_short$responder_mean, k = K_outer, list = TRUE, returnTrain = TRUE)

metrics_nested_short <- vector("list", K_outer)
imp_nested_short     <- vector("list", K_outer)

## ---- 5) Nested CV loop ----
for(k in seq_len(K_outer)){
  tr <- dat_short[ outer_folds[[k]], ]
  te <- dat_short[-outer_folds[[k]], ]

  # Pick inner control here: ctrl_inner5 OR ctrl_inner3
  fit_k <- train(
    responder_mean ~ .,
    data = tr,
    method = "rf",
    metric = "ROC",
    trControl = ctrl_inner5,          # <-- swap to ctrl_inner3 if desired
    tuneGrid = rf_grid_short,
    ntree = rf_ntree,
    importance = TRUE
  )

  # Predict on outer test fold
  prob <- predict(fit_k, newdata = te, type = "prob")[, "Responder"]
  pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                 levels = c("Responder", "NonResponder"))

  # Metrics on outer test fold
  roc_k <- pROC::roc(response = te$responder_mean,
                     predictor = prob,
                     levels = c("NonResponder","Responder"),
                     direction = "<")
  cm_k  <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")

  metrics_nested_short[[k]] <- tibble(
    fold       = k,
    AUC        = as.numeric(pROC::auc(roc_k)),
    Accuracy   = as.numeric(cm_k$overall["Accuracy"]),
    Sensitivity= as.numeric(cm_k$byClass["Sensitivity"]),
    Specificity= as.numeric(cm_k$byClass["Specificity"]),
    Best_mtry  = fit_k$bestTune$mtry,
    n_test     = nrow(te)
  )

  imp_nested_short[[k]] <- get_vi(fit_k, k)
}

nested_perf_short <- bind_rows(metrics_nested_short)
nested_vi_short   <- bind_rows(imp_nested_short)

## ---- 6) Summaries: mean ± SD across outer folds ----
nested_summary_short <- nested_perf_short %>%
  summarise(
    AUC_mean = mean(AUC),            AUC_sd = sd(AUC),
    Acc_mean = mean(Accuracy),       Acc_sd = sd(Accuracy),
    Sens_mean= mean(Sensitivity),    Sens_sd = sd(Sensitivity),
    Spec_mean= mean(Specificity),    Spec_sd = sd(Specificity)
  )

perf_nested_short_table <- tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity"),
  Value  = c(
    sprintf("%.2f ± %.2f", nested_summary_short$AUC_mean, nested_summary_short$AUC_sd),
    sprintf("%.2f ± %.2f", nested_summary_short$Acc_mean, nested_summary_short$Acc_sd),
    sprintf("%.2f ± %.2f", nested_summary_short$Sens_mean, nested_summary_short$Sens_sd),
    sprintf("%.2f ± %.2f", nested_summary_short$Spec_mean, nested_summary_short$Spec_sd)
  )
)

print(perf_nested_short_table)

## ---- 7) Feature stability across outer folds (rank-based) ----
rankstab_nested_short <- nested_vi_short %>%
  group_by(fold) %>%
  arrange(desc(Importance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  group_by(Feature) %>%
  summarise(
    MedianRank = median(rank),
    IQR_L      = quantile(rank, 0.25),
    IQR_U      = quantile(rank, 0.75),
    .groups="drop"
  ) %>%
  arrange(MedianRank)

# Mean ± SD importance across outer folds
imp_nested_summary <- nested_vi_short %>%
  group_by(Feature) %>%
  summarise(
    MeanImportance = mean(Importance, na.rm = TRUE),
    SDImportance   = sd(Importance, na.rm = TRUE),
    .groups="drop"
  )

features_nested_short_table <- left_join(imp_nested_summary, rankstab_nested_short, by = "Feature") %>%
  mutate(
    `Importance (mean ± SD)` = sprintf("%.2f ± %.2f", MeanImportance, SDImportance),
    `Rank (median [IQR])`    = sprintf("%.0f [%.0f–%.0f]", MedianRank, IQR_L, IQR_U)
  ) %>%
  select(Feature, `Importance (mean ± SD)`, `Rank (median [IQR])`) %>%
  arrange(as.numeric(gsub("[^0-9]", "", `Rank (median [IQR])`)))  # nice order

print(features_nested_short_table)

## (Optional) also inspect per-fold chosen mtry
table(nested_perf_short$Best_mtry)
```













## investigation of responder in LEANER
```{r}
# add responder status to meta data
responder_inf <- class.data.cluster %>%
  select(Probanden_ID, responder_mean)

## add information to meta data
meta.leaner <- data.frame(sample_data(PS.tree))
meta.leaner.r <- meta.leaner %>%
  left_join(responder_inf, by = "Probanden_ID") %>%
  filter(!is.na(responder_mean)) %>%
  rename(responder = responder_mean)

meta.leaner.r$responder <- factor(meta.leaner.r$responder, 
                                  levels = c(0, 1),
                                  labels = c("non-responder", "responder"))

########################
####### Plot bmi
########################
meta.leaner.r %>%
  ggplot(aes(x = Visite, y = BMI, fill = responder)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "BMI (kg/m²)") -> fig.3.bmi.responder
fig.3.bmi.responder

##########
### statistics
##############
#### statistics for chao 1 index
meta.leaner.r %>%
  group_by(Visite) %>%
  summarize(mean = mean(BMI),
            sd = sd(BMI))

complete_meta <- meta.leaner.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = BMI ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = BMI ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = BMI ~ responder, 
    paired = FALSE,
    p.adjust.method = "holm")


######################
###### body fat
####################
sample.data.all <- readRDS("rowdata/sample.data.all.rds")
fat.r <- sample.data.all %>%
  left_join(responder_inf, by = "Probanden_ID") %>%
  filter(!is.na(responder_mean)) %>%
  rename(responder = responder_mean) %>%
  filter(Visite != "baseline 0")

fat.r$responder <- factor(fat.r$responder, 
                                  levels = c(0, 1),
                                  labels = c("non-responder", "responder"))

fat.r %>%
  ggplot(aes(x = Visite, y = koerperfett, fill = responder)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "bodyfat (%)") -> fig.3.bodyfat.responder
fig.3.bodyfat.responder

##########
### statistics
##############
#### statistics for chao 1 index
fat.r %>%
  group_by(Visite) %>%
  dplyr::summarize(mean = mean(koerperfett),
            sd = sd(koerperfett))

complete_meta <- fat.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta$Visite <- factor(complete_meta$Visite, 
                                  levels = c("baseline 1", "fasting", "follow-up"))

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = koerperfett ~ responder, 
    paired = FALSE,
    p.adjust.method = "holm")

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = koerperfett ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = koerperfett ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")
```

## changes in the selected features
```{r}
###########
### ldl
########
meta.leaner.r %>%
  ggplot(aes(x = Visite, y = ldl, fill = responder)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "LDL (mg/dL)") -> fig.3.ldl.responder
fig.3.ldl.responder

##########
### statistics
##############
#### statistics for chao 1 index
meta.leaner.r %>%
  group_by(Visite) %>%
  summarize(mean = mean(ldl),
            sd = sd(ldl))

complete_meta <- meta.leaner.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = ldl ~ responder, 
    paired = FALSE)

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = ldl ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = ldl ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

complete_meta %>% 
  group_by(responder) %>%
  pairwise_wilcox_test(
    formula = ldl ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")



###########
### systolic blood pressure
########
meta.leaner.r %>%
  ggplot(aes(x = Visite, y = sys_mean, fill = responder)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "sys BP (mmHg)") -> fig.3.sysBP.responder
fig.3.sysBP.responder

##########
### statistics
##############
#### statistics for chao 1 index
meta.leaner.r %>%
  group_by(Visite) %>%
  summarize(mean = mean(sys_mean),
            sd = sd(sys_mean))

complete_meta <- meta.leaner.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = sys_mean ~ responder, 
    paired = FALSE)

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = sys_mean ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = sys_mean ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######################
###### oscillibacter
#######################
all.r <- class_data_all %>%
  left_join(responder_inf, by = "Probanden_ID") %>%
  filter(!is.na(responder_mean)) %>%
  rename(responder = responder_mean)

all.r$responder <- factor(all.r$responder,
                                       levels = c(0, 1), 
                                       labels = c("non-responder", "responder"))

all.r %>%
  ggplot(aes(x = Visite, y = Oscillibacter.sp..57_20., fill = responder)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Oscilli (counts)") -> fig.3.Oscilli.responder
fig.3.Oscilli.responder

##########
### statistics
##############
#### statistics for chao 1 index
all.r %>%
  group_by(Visite) %>%
  summarize(mean = mean(Oscillibacter.sp..57_20.),
            sd = sd(Oscillibacter.sp..57_20.))

complete_meta <- all.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ responder, 
    paired = FALSE)

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = Oscillibacter.sp..57_20. ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######################
###### Faecalibacterium
#######################
all.r %>%
  ggplot(aes(x = Visite, y = NA.Faecalibacterium.species.incertae.sedis, fill = responder)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("non-responder" = "#fc8d59", "responder" = "#998ec3")) +
  geom_point(position = position_jitterdodge(jitter.width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  theme(legend.position = "top",
        legend.title = element_blank())+
  labs(y = "Faecali (counts)") -> fig.3.Faecali.responder
fig.3.Faecali.responder

##########
### statistics
##############
#### statistics for chao 1 index
all.r %>%
  group_by(Visite) %>%
  summarize(mean = mean(NA.Faecalibacterium.species.incertae.sedis),
            sd = sd(NA.Faecalibacterium.species.incertae.sedis))

complete_meta <- all.r %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_meta %>% 
  group_by(Visite) %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ responder, 
    paired = FALSE)

######
## non-responder
#############
complete_meta %>% 
  filter(responder == "non-responder") %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

######
## responder
#############
complete_meta %>% 
  filter(responder == "responder") %>%
  pairwise_wilcox_test(
    formula = NA.Faecalibacterium.species.incertae.sedis ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")


```
## Correlation of features
```{r}
library(corrplot)
library(Hmisc)
delta.bmi <- bmi.total %>%
  select(Probanden_ID, dif)

class.data.new <- class.data.cluster %>%
  left_join(delta.bmi, by = "Probanden_ID") %>%
  rename("delta_BMI" = "dif")

class.data.cluster.50 <- class.data.new

# select feature for correlation matrix
cor_features <- class.data.new %>% 
  select(Oscillibacter.sp..57_20.,
         NA.Faecalibacterium.species.incertae.sedis,
         sys_mean,
         ldl,
         BMI) %>%
  na.omit()


cor_names <- cor_features
colnames(cor_features) <- c("Oscilli sp.\n57_20", 
  "Faecali sp.\nincertae sedis", 
  "Sys BP", 
  "LDL", 
  "BMI"
)

cor_test_result <- rcorr(as.matrix(cor_features, type = "spearman"))  
#the data might not be normal distributed

print(cor_test_result$r)  
print(cor_test_result$P) 
cor_test_result$P[is.na(cor_test_result$P)] <- 1 #

# Extract the correlation matrix (r) and p-value matrix (P)
cor_matrix <- cor_test_result$r  
p_matrix <- cor_test_result$P    


# Plot the correlation matrix with significant correlations highlighted
corrplot(cor_matrix, 
         type = "upper",        
         order = "hclust",       
         p.mat = p_matrix,       
         sig.level = 0.05,       
         insig = "blank")        

#svg("fig.3.corrplot.short.svg", width = 8, height = 8)
corrplot(cor_matrix, 
         type = "upper",         
         order = "hclust",       
         p.mat = p_matrix,       
         sig.level = 0.05,       
         insig = "blank",        
         tl.col = "black",       
         tl.srt = 45,            
         tl.cex = 0.8,           
         cl.cex = 0.8,           
         addCoef.col = "black",  
         number.cex = 0.7,       
         col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(100), 
         mar = c(0, 0, 1, 0))    
dev.off()

# other plot
#svg("fig.3.corrplot.correlation.non.signifcant.svg", width = 8, height = 8)
corrplot(cor_matrix, type = "upper", order = "hclust", 
         tl.col = "black", tl.cex = 1.5, cl.cex = 1.3,
         number.cex = 1.5)
dev.off()
```


```{r}

library(ggcorrplot)
cor_matrix <- cor_test_result$r
p_matrix <- cor_test_result$P
p_matrix[is.na(p_matrix)] <- 1 

# Plot
fig.3.LEANER.heatmap <- ggcorrplot(cor_matrix,
           method = "square",        
           type = "upper",           
           lab = TRUE,               # Show correlation numbers
           lab_size = 6,             # Correlation number size (~24 pt)
           colors = c("#0075DC", "white", "#E64B35FF"),  
           tl.cex = 1.5,             # Axis label size (~24 pt)
           legend.title = "Corr",    # Optional: set a custom legend title
           ggtheme = theme_minimal(base_size = 24)) +  # Base text size 24
  theme(legend.text = element_text(size = 20),
        legend.title = element_text(size = 24),
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        plot.title = element_text(size = 24))

fig.3.LEANER.heatmap
```


## Compare model containing clinical features and BMI with model including microbial features
Does the microbiome add extra predictive information

```{r}
# ===============================
# Compare short model (4 feats) vs. no-LDL (3 feats)
# Same 100 resamples, inner 5-fold tuning, paired comparison
# ===============================

library(tidyverse)
library(caret)
library(pROC)
library(randomForest)

set.seed(2025)

# ---- 0) Prepare data (same encoding as before) ----
# Adjust names if needed to match your data frame
short_features <- c("Oscillibacter.sp..57_20.",
                    "NA.Faecalibacterium.species.incertae.sedis",
                    "ldl", "sys_mean", "BMI")

dat <- df %>%
  mutate(
    responder_mean = factor(
      ifelse(as.character(responder_mean) %in% c("1", 1, "Responder"),
             "Responder", "NonResponder"),
      levels = c("Responder", "NonResponder")
    )
  ) %>%
  select(all_of(c(short_features, "responder_mean"))) %>%
  drop_na()

# Define the 3-feature set (drop ldl)
feat_clinical  <- c("ldl", "sys_mean", "BMI")

feat_microbiome   <- c("Oscillibacter.sp..57_20.",
                    "NA.Faecalibacterium.species.incertae.sedis",
                    "sys_mean", "ldl", "BMI")

# ---- 1) Common training control (inner 5-fold CV, optimize ROC) ----
ctrl_inner <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

rf_ntree <- 1000

# Helper: make a tuning grid for a given #predictors p
mk_grid <- function(p){
  expand.grid(mtry = seq(1, min(p, floor(sqrt(p)) + 2), by = 1))
}

# ---- 2) Precompute SAME 100 stratified 80/20 splits (paired) ----
B <- 100
set.seed(4242)  # controls outer splits
split_index <- vector("list", B)
for(b in seq_len(B)){
  split_index[[b]] <- createDataPartition(dat$responder_mean, p = 0.8, list = FALSE)
}

# ---- 3) Function to run the repeated CV for a given feature set ----
run_repeated <- function(dat, feature_names, split_index, ctrl_inner, rf_ntree){
  p <- length(feature_names)
  rf_grid <- mk_grid(p)

  metrics_list <- vector("list", length(split_index))

  for(b in seq_along(split_index)){
    tr_idx <- split_index[[b]]
    tr <- dat[tr_idx, c(feature_names, "responder_mean")]
    te <- dat[-tr_idx, c(feature_names, "responder_mean")]

    fit_b <- train(
      responder_mean ~ .,
      data = tr,
      method = "rf",
      metric = "ROC",
      trControl = ctrl_inner,
      tuneGrid = rf_grid,
      ntree = rf_ntree,
      importance = TRUE
    )

    # Predict on held-out 20%
    prob <- predict(fit_b, newdata = te, type = "prob")[, "Responder"]
    pred <- factor(ifelse(prob >= 0.5, "Responder", "NonResponder"),
                   levels = c("Responder","NonResponder"))

    roc_b <- pROC::roc(response = te$responder_mean,
                       predictor = prob,
                       levels = c("NonResponder","Responder"),
                       direction = "<")

    cm <- caret::confusionMatrix(pred, te$responder_mean, positive = "Responder")

    # Balanced accuracy, F1, MCC from confusion matrix parts
    TP <- cm$table["Responder","Responder"]
    TN <- cm$table["NonResponder","NonResponder"]
    FP <- cm$table["Responder","NonResponder"]
    FN <- cm$table["NonResponder","Responder"]

    sens <- as.numeric(cm$byClass["Sensitivity"])
    spec <- as.numeric(cm$byClass["Specificity"])
    balacc <- mean(c(sens, spec), na.rm = TRUE)

    precision <- ifelse((TP+FP) > 0, TP/(TP+FP), NA_real_)
    recall    <- sens
    F1 <- ifelse(is.na(precision) | is.na(recall) | (precision+recall)==0,
                 NA_real_, 2*precision*recall/(precision+recall))

    denom <- sqrt((TP+FP)*(TP+FN)*(TN+FP)*(TN+FN))
    MCC <- ifelse(denom > 0, ((TP*TN)-(FP*FN))/denom, NA_real_)

    metrics_list[[b]] <- tibble(
      iter       = b,
      AUC        = as.numeric(pROC::auc(roc_b)),
      Accuracy   = as.numeric(cm$overall["Accuracy"]),
      Sensitivity= sens,
      Specificity= spec,
      BalancedAccuracy = balacc,
      F1         = F1,
      MCC        = MCC,
      Best_mtry  = fit_b$bestTune$mtry
    )
  }

  bind_rows(metrics_list)
}

# ---- 4) Run both models on identical splits ----
mc_clinical  <- run_repeated(dat, feat_clinical, split_index, ctrl_inner, rf_ntree)
mc_microbiome    <- run_repeated(dat, feat_microbiome, split_index, ctrl_inner, rf_ntree)

# ---- 5) Summaries (mean±SD and median [IQR]) for each ----
summarize_perf <- function(df){
  tibble(
    AUC_mean = mean(df$AUC, na.rm=TRUE), AUC_sd = sd(df$AUC, na.rm=TRUE),
    Acc_mean = mean(df$Accuracy, na.rm=TRUE), Acc_sd = sd(df$Accuracy, na.rm=TRUE),
    Sens_mean= mean(df$Sensitivity, na.rm=TRUE), Sens_sd = sd(df$Sensitivity, na.rm=TRUE),
    Spec_mean= mean(df$Specificity, na.rm=TRUE), Spec_sd = sd(df$Specificity, na.rm=TRUE),
    BalAcc_mean = mean(df$BalancedAccuracy, na.rm=TRUE), BalAcc_sd = sd(df$BalancedAccuracy, na.rm=TRUE),
    F1_mean = mean(df$F1, na.rm=TRUE), F1_sd = sd(df$F1, na.rm=TRUE),
    MCC_mean = mean(df$MCC, na.rm=TRUE), MCC_sd = sd(df$MCC, na.rm=TRUE),
    AUC_median = median(df$AUC, na.rm=TRUE),
    AUC_IQR_L  = quantile(df$AUC, 0.25, na.rm=TRUE),
    AUC_IQR_U  = quantile(df$AUC, 0.75, na.rm=TRUE)
  )
}

sum_clinical <- summarize_perf(mc_clinical)
sum_microbiome   <- summarize_perf(mc_microbiome)

# Nicely formatted tables (mean ± SD)
fmt_ms <- function(m, s) sprintf("%.2f ± %.2f", m, s)

perf_microbiome <- tibble::tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity","Balanced Accuracy","F1","MCC"),
  Value  = c(
    fmt_ms(sum_microbiome$AUC_mean, sum_microbiome$AUC_sd),
    fmt_ms(sum_microbiome$Acc_mean, sum_microbiome$Acc_sd),
    fmt_ms(sum_microbiome$Sens_mean, sum_microbiome$Sens_sd),
    fmt_ms(sum_microbiome$Spec_mean, sum_microbiome$Spec_sd),
    fmt_ms(sum_microbiome$BalAcc_mean, sum_microbiome$BalAcc_sd),
    fmt_ms(sum_microbiome$F1_mean, sum_microbiome$F1_sd),
    fmt_ms(sum_microbiome$MCC_mean, sum_microbiome$MCC_sd)
  ),
  Model = "Model Microbiome"
)

perf_clinical <- tibble::tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity","Balanced Accuracy","F1","MCC"),
  Value  = c(
    fmt_ms(sum_clinical$AUC_mean, sum_clinical$AUC_sd),
    fmt_ms(sum_clinical$Acc_mean, sum_clinical$Acc_sd),
    fmt_ms(sum_clinical$Sens_mean, sum_clinical$Sens_sd),
    fmt_ms(sum_clinical$Spec_mean, sum_clinical$Spec_sd),
    fmt_ms(sum_clinical$BalAcc_mean, sum_clinical$BalAcc_sd),
    fmt_ms(sum_clinical$F1_mean, sum_clinical$F1_sd),
    fmt_ms(sum_clinical$MCC_mean, sum_clinical$MCC_sd)
  ),
  Model = "Model Clinical"
)
perf_microbiome
perf_clinical

bind_rows(perf_microbiome, perf_clinical)

# ---- 6) Paired comparison: deltas per split (3 feats minus 4 feats) ----
cmp <- mc_microbiome %>%
  select(iter, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy, F1, MCC) %>%
  rename_with(~ paste0(.x, "_microbiome"), -iter) %>%
  inner_join(
    mc_clinical %>%
      select(iter, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy, F1, MCC) %>%
      rename_with(~ paste0(.x, "_clinical"), -iter),
    by = "iter"
  ) %>%
  mutate(
    d_AUC  = AUC_microbiome  - AUC_clinical,
    d_Acc  = Accuracy_microbiome - Accuracy_clinical,
    d_Sens = Sensitivity_microbiome - Sensitivity_clinical,
    d_Spec = Specificity_microbiome - Specificity_clinical,
    d_BalAcc = BalancedAccuracy_microbiome - BalancedAccuracy_clinical,
    d_F1   = F1_microbiome - F1_clinical,
    d_MCC  = MCC_microbiome - MCC_clinical
  )

# Summary of differences (median [IQR] and mean ± SD)
# helper to format quantiles with NA handling
q25 <- function(x) quantile(x, 0.25, na.rm = TRUE)
q75 <- function(x) quantile(x, 0.75, na.rm = TRUE)

delta_summary <- tibble(
  Metric = c("AUC","Accuracy","Sensitivity","Specificity","Balanced Accuracy","F1","MCC"),
  MeanDiff   = c(mean(cmp$d_AUC,  na.rm=TRUE),
                 mean(cmp$d_Acc,  na.rm=TRUE),
                 mean(cmp$d_Sens, na.rm=TRUE),
                 mean(cmp$d_Spec, na.rm=TRUE),
                 mean(cmp$d_BalAcc, na.rm=TRUE),
                 mean(cmp$d_F1,   na.rm=TRUE),
                 mean(cmp$d_MCC,  na.rm=TRUE)),
  SDDiff     = c(sd(cmp$d_AUC,  na.rm=TRUE),
                 sd(cmp$d_Acc,  na.rm=TRUE),
                 sd(cmp$d_Sens, na.rm=TRUE),
                 sd(cmp$d_Spec, na.rm=TRUE),
                 sd(cmp$d_BalAcc, na.rm=TRUE),
                 sd(cmp$d_F1,   na.rm=TRUE),
                 sd(cmp$d_MCC,  na.rm=TRUE)),
  MedianDiff = c(median(cmp$d_AUC,  na.rm=TRUE),
                 median(cmp$d_Acc,  na.rm=TRUE),
                 median(cmp$d_Sens, na.rm=TRUE),
                 median(cmp$d_Spec, na.rm=TRUE),
                 median(cmp$d_BalAcc, na.rm=TRUE),
                 median(cmp$d_F1,   na.rm=TRUE),
                 median(cmp$d_MCC,  na.rm=TRUE)),
  IQR_L      = c(q25(cmp$d_AUC),  q25(cmp$d_Acc),  q25(cmp$d_Sens),
                 q25(cmp$d_Spec), q25(cmp$d_BalAcc), q25(cmp$d_F1), q25(cmp$d_MCC)),
  IQR_U      = c(q75(cmp$d_AUC),  q75(cmp$d_Acc),  q75(cmp$d_Sens),
                 q75(cmp$d_Spec), q75(cmp$d_BalAcc), q75(cmp$d_F1), q75(cmp$d_MCC)),
  N_used     = c(sum(!is.na(cmp$d_AUC)),
                 sum(!is.na(cmp$d_Acc)),
                 sum(!is.na(cmp$d_Sens)),
                 sum(!is.na(cmp$d_Spec)),
                 sum(!is.na(cmp$d_BalAcc)),
                 sum(!is.na(cmp$d_F1)),
                 sum(!is.na(cmp$d_MCC)))
) %>%
  mutate(
    `Mean ± SD`    = sprintf("%.03f ± %.03f", MeanDiff, SDDiff),
    `Median [IQR]` = sprintf("%.03f [%.03f–%.03f]", MedianDiff, IQR_L, IQR_U)
  ) %>%
  select(Metric, `Mean ± SD`, `Median [IQR]`, N_used)


########################
### the microbiome adds extra positive value on the model even if we include BMI at baseline
delta_summary
```
