---
title: "Fig.1 - Clinical data"
author: "Gelsomina Nicola Kaufhold"
date: "2025-06-18"
output: html_document
---

# Figure 2: Proof of concept that fasting worked

# Preperation

## Load required packages

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(phyloseq)
library(tidyverse)
library(readxl)
library(ggrain)
library(rstatix)
library(readxl)
library(dplyr)
```


## Adjust themes for plotting

```{r warning=FALSE}
#ggsave("Fig.3.V1.V2.comparison.beta.diversity.pdf", plot = Fig.3.V1.V2.comparison.beta.diversity, width = 5, height = 5)
boxplot.theme.legend <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black",angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "top")

boxplot.theme <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")

boxplot.theme.2 <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")


########################################
### setting colors
#########################################

color.scale.4 <- scale_fill_manual(values = c(
  "baseline 0" = "#FFD700",  # Baseline 0
  "baseline" = "#E69F00",    # Baseline 1
  "fasting" = "#CC79A7",     # Fasting
  "follow-up" = "#0072B2"    # Follow-up
))

color.scale.3 <- scale_fill_manual(values = c(
  "baseline" = "#E69F00",    # Baseline 1
  "fasting" = "#CC79A7",     # Fasting
  "follow-up" = "#0072B2"    # Follow-up
))

color.fill.sex <- scale_fill_manual(values = c(
  "male" = "#998ec3",    
  "female" = "#5ab4ac"))

```

# Metadata: Physical changes to fasting


## Prepare data

```{r warning=FALSE}
#### Prepare data
sample.data <- read_excel("rowdata/raw_data_redcap_HB.xlsx")
sample.data <- sample.data %>%
  rename(`Probanden_ID`= "Probanden-ID")

sample.data$timepoint <- sample.data$Visite

sample.data$Visite <- factor(sample.data$Visite,
                              levels = c("1", "2", "3", "4"),
                              labels = c("baseline 0", "baseline 1", "fasting", "follow-up"))


## ad weight to meta data to calculate BMI
hight <- read_excel("rowdata/weight.xlsx", 
    col_names = FALSE)

hight <- hight[ , c(7, 11,9)]
colnames(hight) <- c("proband_id", "Hight", "Sex")

meta.all <- sample.data %>%
  left_join(hight, by = c("Probanden_ID" = "proband_id"))

meta.all.bmi <- meta.all %>%
  dplyr::mutate(Hight = Hight / 100) %>%
  mutate(BMI = koerpergewicht / (Hight^2))

rownames(meta.all.bmi) <- meta.all.bmi$Novogene_ID

meta.all.bmi$BMI <- as.numeric(meta.all.bmi$BMI)

sample.data.all <- meta.all.bmi


sample.data.all$Sex <- factor(sample.data.all$Sex,
                              levels = c("w", "m"),
                              labels = c("female", "male"))
#saveRDS(sample.data.all, "rowdata/sample.data.all.rds")

```

## Physical changes to fasting
raincloud plots by: Allen, M., Poggiali, D., Whitaker, K., Marshall, T. R., van Langen, J., & Kievit, R. A. Raincloud plots: a multi-platform tool for robust data visualization Wellcome Open Research 2021, 4:63. https://doi.org/10.12688/wellcomeopenres.15191.2


```{r echo=TRUE, warning=FALSE}
#################
## Changes in BMI
#################
sample.data.all %>%
  ggplot(aes(x = Visite, y = BMI, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "BMI") ->  bmi.fasting.plot
 bmi.fasting.plot

##################
###### statistics
##################
 
 
###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(BMI),
            sd = sd(BMI))

#### significance

# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = BMI ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = BMI ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_BMI <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_BMI


BMI_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_BMI = mean(BMI, na.rm = TRUE),
    sd_BMI = sd(BMI, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_BMI <- combined_results_BMI %>%
  left_join(BMI_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_BMI,
    sd_group1 = sd_BMI
  )

combined_results_BMI <- combined_results_BMI %>%
  left_join(BMI_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_BMI,
    sd_group2 = sd_BMI
  )

####################################
## Changes in systolic blood pressure
#####################################
sample.data.all %>%
  ggplot(aes(x = Visite, y = sys_mean, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Sys BP") -> bp.sys.fasting.plot
bp.sys.fasting.plot

######################
####### statistics
####################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(sys_mean),
            sd = sd(sys_mean))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = sys_mean ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = sys_mean ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_sysbp <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_sysbp

sysbp_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_sysbp = mean(sys_mean, na.rm = TRUE),
    sd_sysbp = sd(sys_mean, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_sysbp <- combined_results_sysbp %>%
  left_join(sysbp_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_sysbp,
    sd_group1 = sd_sysbp
  )

combined_results_sysbp <- combined_results_sysbp %>%
  left_join(sysbp_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_sysbp,
    sd_group2 = sd_sysbp
  )

# View combined results
combined_results_sysbp

###############
## body fat
##############
sample.data.all %>%
  ggplot(aes(x = Visite, y = koerperfett, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Body fat in %") -> koerperfett.fasting.plot
koerperfett.fasting.plot


######################
####### statistics
######################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(koerperfett),
            sd = sd(koerperfett))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = koerperfett ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = koerperfett ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_bodyfat <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_bodyfat


bodyfat_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_bodyfat = mean(koerperfett, na.rm = TRUE),
    sd_bodyfat = sd(koerperfett, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_bodyfat <- combined_results_bodyfat %>%
  left_join(bodyfat_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_bodyfat,
    sd_group1 = sd_bodyfat
  )

combined_results_bodyfat <- combined_results_bodyfat %>%
  left_join(bodyfat_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_bodyfat,
    sd_group2 = sd_bodyfat
  )

# View combined results
combined_results_bodyfat


#########################
##### bodyfat in absolute values
#########################
sample.data.all$bodyfat_absolute <- sample.data.all$koerpergewicht*(sample.data.all$koerperfett/100)

sample.data.all %>%
  ggplot(aes(x = Visite, y = bodyfat_absolute, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Body fat in kg") -> bodyfat_absolute.fasting.plot
bodyfat_absolute.fasting.plot


######################
####### statistics
######################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(bodyfat_absolute),
            sd = sd(bodyfat_absolute))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = bodyfat_absolute ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = bodyfat_absolute ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_bodyfat_absolute <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_bodyfat_absolute


bodyfat_absolute_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_bodyfat = mean(bodyfat_absolute, na.rm = TRUE),
    sd_bodyfat = sd(bodyfat_absolute, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_bodyfat_absolute <- combined_results_bodyfat_absolute %>%
  left_join(bodyfat_absolute_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_bodyfat,
    sd_group1 = sd_bodyfat
  )

combined_results_bodyfat_absolute <- combined_results_bodyfat_absolute %>%
  left_join(bodyfat_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_bodyfat,
    sd_group2 = sd_bodyfat
  )

# View combined results
combined_results_bodyfat_absolute






###############
## lean mass
#############
sample.data.all %>%
  ggplot(aes(x = Visite, y = fettfreie_masse, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Lean mass in %") -> leanmass.fasting.plot
leanmass.fasting.plot

#########################
### statistics
####################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(fettfreie_masse),
            sd = sd(fettfreie_masse))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = fettfreie_masse ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = fettfreie_masse ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_leanmass <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_leanmass

leanmass_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_leanmass = mean(fettfreie_masse, na.rm = TRUE),
    sd_leanmass = sd(fettfreie_masse, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_leanmass <- combined_results_leanmass %>%
  left_join(leanmass_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_leanmass,
    sd_group1 = sd_leanmass
  )

combined_results_leanmass <- combined_results_leanmass %>%
  left_join(leanmass_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_leanmass,
    sd_group2 = sd_leanmass
  )

# View combined results
combined_results_leanmass

###########################
######## asbolute leanmass
#############################
sample.data.all$leanmass_absolute <- sample.data.all$koerpergewicht*(sample.data.all$fettfreie_masse/100)

sample.data.all %>%
  ggplot(aes(x = Visite, y = leanmass_absolute, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Leanmass in kg") -> leanmass_absolute.fasting.plot
leanmass_absolute.fasting.plot


######################
####### statistics
######################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(leanmass_absolute),
            sd = sd(leanmass_absolute))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = leanmass_absolute ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = leanmass_absolute ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_leanmass_absolute <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_leanmass_absolute


leanmass_absolute_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_leanmass_absolute = mean(leanmass_absolute, na.rm = TRUE),
    sd_leanmass_absolute = sd(leanmass_absolute, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_leanmass_absolute <- combined_results_leanmass_absolute %>%
  left_join(leanmass_absolute_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_leanmass_absolute,
    sd_group1 = sd_leanmass_absolute
  )

combined_results_leanmass_absolute <- combined_results_leanmass_absolute %>%
  left_join(leanmass_absolute_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_leanmass_absolute,
    sd_group2 = sd_leanmass_absolute
  )

# View combined results
combined_results_leanmass_absolute


############################
######### WHR
##########################

sample.data.all %>%
  ggplot(aes(x = Visite, y = WHR, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "WHR") -> WHR.fasting.plot
WHR.fasting.plot
#########################
### statistics
####################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(WHR),
            sd = sd(WHR))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = WHR ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = WHR ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_WHR <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))
combined_results_WHR

WHR_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_WHR = mean(WHR, na.rm = TRUE),
    sd_WHR = sd(WHR, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_WHR <- combined_results_WHR %>%
  left_join(WHR_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_WHR,
    sd_group1 = sd_WHR
  )

combined_results_WHR <- combined_results_WHR %>%
  left_join(WHR_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_WHR,
    sd_group2 = sd_WHR
  )

# View combined results
combined_results_WHR



####################
######## diastolic blood pressure
##################

sample.data.all %>%
  ggplot(aes(x = Visite, y = bp_dia_mean, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Dia BP") -> bp.dia.mean.fasting.plot
bp.dia.mean.fasting.plot

#########################
### statistics
####################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(bp_dia_mean),
            sd = sd(bp_dia_mean))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = bp_dia_mean ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = bp_dia_mean ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_diabp <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))
combined_results_diabp

diabp_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_diabp = mean(bp_dia_mean, na.rm = TRUE),
    sd_diabp = sd(bp_dia_mean, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_diabp <- combined_results_diabp %>%
  left_join(diabp_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_diabp,
    sd_group1 = sd_diabp
  )

combined_results_diabp <- combined_results_diabp %>%
  left_join(diabp_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_diabp,
    sd_group2 = sd_diabp
  )

# View combined results
combined_results_diabp


####################
######## Heart rate
##################

sample.data.all %>%
  ggplot(aes(x = Visite, y = bp_bpm_mean, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "Heart rate") -> heartrate.fasting.plot
heartrate.fasting.plot

#########################
### statistics
####################

###### mean, sd
sample.data.all %>%
  group_by(Visite) %>%
  summarize(mean = mean(bp_bpm_mean),
            sd = sd(bp_bpm_mean))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- sample.data.all %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = bp_bpm_mean ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- sample.data.all %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = bp_bpm_mean ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_hr <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))
combined_results_hr

hr_summary <- sample.data.all %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_hr = mean(bp_bpm_mean, na.rm = TRUE),
    sd_hr = sd(bp_bpm_mean, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_hr <- combined_results_hr %>%
  left_join(hr_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_hr,
    sd_group1 = sd_hr
  )

combined_results_hr <- combined_results_hr %>%
  left_join(hr_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_hr,
    sd_group2 = sd_hr
  )

# View combined results
combined_results_hr

```
## Changes in REE and RER
```{r}
##########################
### Changes in REE
#######################
REE <- sample.data.all %>%
  select(Probanden_ID, Sex, Visite, rnu_kcal_d)

###### boxplots
REE %>%
  ggplot(aes(x = Visite, y = rnu_kcal_d, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  color.scale.4 +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  boxplot.theme +
  labs(y = "REE") -> fig.1.REE
fig.1.REE

#### violine plots
REE %>%
ggplot(aes(x = Visite, y = rnu_kcal_d, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "REE") -> fig.1.REE.violine
fig.1.REE.violine

###################
##### statistics
##################

###### mean, sd
REE %>%
  group_by(Visite) %>%
  summarize(mean = mean(rnu_kcal_d),
            sd = sd(rnu_kcal_d))

#### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- REE %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = rnu_kcal_d ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- REE %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = rnu_kcal_d ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_REE <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_REE

REE_summary <- REE %>%
  dplyr::group_by(Visite) %>%
  summarize(
    mean_REE = mean(rnu_kcal_d, na.rm = TRUE),
    sd_REE = sd(rnu_kcal_d, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_REE <- combined_results_REE %>%
  left_join(REE_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_REE,
    sd_group1 = sd_REE
  )

combined_results_REE <- combined_results_REE %>%
  left_join(REE_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_REE,
    sd_group2 = sd_REE
  )


##########################
### Changes in RER
#######################
RER <- sample.data.all %>%
  select(Probanden_ID, Sex, Visite, rer)
  
RER %>%
  ggplot(aes(x = Visite, y = rer, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  color.scale.4 +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  boxplot.theme +
  labs(y = "RER") -> fig.1.RER
fig.1.RER

#### violine plots

RER %>%
ggplot(aes(x = Visite, y = rer, fill = Sex)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
    geom_rain(alpha = 0.5,
              shape = 21,
              point.args = list(size = 3, shape = 21),
              boxplot.args = list(fill = "gray80", color = "black", outlier.shape = NA)) +
  color.fill.sex +
  boxplot.theme.legend +
  labs(y = "RER") -> fig.1.RER.violine

####################
####### statistics
###################

###### mean, sd
RER <- data.frame(RER)
RER %>%
  group_by(Visite) %>%
  summarize(mean = mean(rer),
            sd = sd(rer))

### significance
# ---- Part 1: Compare Group 1 vs. Group 2 ----
# Filter to subjects that have measurements in both group 1 and group 2
data_1_2 <- RER %>%
  filter(Visite %in% c("baseline 0", "baseline 1")) %>%
  group_by(Probanden_ID) %>%
  filter(n_distinct(Visite) == 2) %>%
   mutate(Visite = factor(Visite, levels = c("baseline 0", "baseline 1"))) %>%
  ungroup()

# Perform the paired Wilcoxon test for group 1 vs 2
result_1_2 <- data_1_2 %>%
  pairwise_wilcox_test(
    formula = rer ~ Visite,
    paired = TRUE,
    p.adjust.method = "none"
  )


# ---- Part 2: Compare Groups 2, 3, and 4 ----
# Filter to subjects with complete data in groups 2, 3, and 4
data_2_3_4 <- RER %>%
  filter(Visite %in% c("baseline 1", "fasting", "follow-up")) %>%
  group_by(Probanden_ID) %>%
  # Keep only subjects with all three visits
  filter(length(unique(Visite)) == 3) %>%
  mutate(Visite = factor(Visite, levels = c("baseline 1", "fasting", "follow-up"))) %>%
  ungroup()

# Perform the paired Wilcoxon tests for all pairwise comparisons among groups 2, 3, and 4
result_2_3_4 <- data_2_3_4 %>%
  pairwise_wilcox_test(
    formula = rer ~ Visite,
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "none"  # Again, postpone adjustment
  )

# ---- Combine and Adjust p-values ----
# Bind the two sets of results and then adjust the combined p-values with holm
combined_results_RER <- bind_rows(result_1_2, result_2_3_4) %>%
  mutate(p.adj = p.adjust(p, method = "holm"))

# View combined results
combined_results_RER

## add mean and sd for each group comparison
rer_summary <- RER %>%
  group_by(Visite) %>%
  summarize(
    mean_rer = mean(rer, na.rm = TRUE),
    sd_rer = sd(rer, na.rm = TRUE),
    .groups = "drop"
  )

combined_results_RER <- combined_results_RER %>%
  left_join(rer_summary, by = c("group1" = "Visite")) %>%
  rename(
    mean_group1 = mean_rer,
    sd_group1 = sd_rer
  )

combined_results_RER <- combined_results_RER %>%
  left_join(rer_summary, by = c("group2" = "Visite")) %>%
  rename(
    mean_group2 = mean_rer,
    sd_group2 = sd_rer
  )

```

## Create table to summarize all results

```{r}
statistics.complete <- rbind(combined_results_BMI, combined_results_bodyfat, combined_results_bodyfat_absolute, combined_results_leanmass, combined_results_leanmass_absolute, combined_results_WHR, combined_results_hr, combined_results_sysbp, combined_results_diabp, combined_results_REE, combined_results_RER)

statistics.complete

## store as xlsx file
library(writexl)
#write_xlsx(statistics.complete, "statistics_results.xlsx")
```

## Over time metadata

### Glucose
```{r}
Glucose <- read_excel("rowdata/meta data/Glucose_data.xlsx")
Glucose_long <- pivot_longer(Glucose, cols = c(4:ncol(Glucose)), names_to = "Probanden_ID", values_to = "Glucose")

df_summary <- Glucose_long %>%
  group_by(Day, Morning_Evening) %>%
  dplyr::summarize(
    mean_glucose = mean(Glucose, na.rm = TRUE),
    sd_glucose   = sd(Glucose, na.rm = TRUE),
    n            = sum(!is.na(Glucose)),
    se_glucose   = sd_glucose / sqrt(n),  # standard error
    .groups      = "drop"
  )

time_levels <- c(
  "Day 1 M", "Day 1 E", 
  "Day 2 M", "Day 2 E",
  "Day 3 M", "Day 3 E",
  "Day 4 M", "Day 4 E",
  "Day 5 M", "Day 5 E",
  "Day 6 M", "Day 6 E",
  "Day 7 M", "Day 7 E"
)

df_summary <- df_summary %>%
  mutate(DayTime = paste0("Day ", Day, " ", Morning_Evening),
         DayTime = factor(DayTime, levels = time_levels))

ggplot(df_summary, aes(x = DayTime, y = mean_glucose, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_glucose - se_glucose,
      ymax = mean_glucose + se_glucose
    ),
    width = 0.2,
    color = "steelblue"
  ) +
  labs(
    x = "Day and time",
    y = "Glucose (mg/dL)"
  ) +
  theme_classic(base_size = 24) +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) -> fig.1.glucose.over.time
fig.1.glucose.over.time

```

### beta-hydroxybutyrate

```{r}
BHB <- read_excel("rowdata/meta data/BHB_data.xlsx")
BHB_long <- pivot_longer(BHB, cols = c(4:ncol(BHB)), names_to = "Probanden_ID", values_to = "BHB")

df_summary <- BHB_long %>%
  group_by(Day, Morning_Evening) %>%
  dplyr::summarize(
    mean_BHB = mean(BHB, na.rm = TRUE),
    sd_BHB   = sd(BHB, na.rm = TRUE),
    n            = sum(!is.na(BHB)),
    se_BHB   = sd_BHB / sqrt(n),  # standard error
    .groups      = "drop"
  )

time_levels <- c(
  "Day 1 M", "Day 1 E", 
  "Day 2 M", "Day 2 E",
  "Day 3 M", "Day 3 E",
  "Day 4 M", "Day 4 E",
  "Day 5 M", "Day 5 E",
  "Day 6 M", "Day 6 E",
  "Day 7 M", "Day 7 E"
)

df_summary <- df_summary %>%
  mutate(DayTime = paste0("Day ", Day, " ", Morning_Evening),
         DayTime = factor(DayTime, levels = time_levels))

ggplot(df_summary, aes(x = DayTime, y = mean_BHB, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_BHB - se_BHB,
      ymax = mean_BHB + se_BHB
    ),
    width = 0.2,
    color = "steelblue"
  ) +
  labs(
    x = "Day and time",
    y = "BHB (mmmol/L)"
  ) +
  theme_classic(base_size = 24) +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) -> fig.1.BHB.over.time
fig.1.BHB.over.time
```

