---
title: "Fig.2 - Microbiome"
author: "Gelsomina Nicola Kaufhold"
date: "2025-06-18"
output: html_document
---

# Figure 2: Microbial changes due fasting

# Preperation

## Loading required packages
```{r message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(rstatix)
library(coin)
library(microbiome)
library(ggpubr)
library(vegan)
library(ggplot2)
library(dplyr)
```

## Load themes and colors
```{r}
boxplot.theme.legend <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black",angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "top")

boxplot.theme <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.title.x = element_blank(),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")

boxplot.theme.2 <- theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none")

########################################
### setting colors
#########################################

color.scale.4 <- scale_fill_manual(values = c(
  "baseline 0" = "#FFD700",  # Baseline 0
  "baseline 1" = "#E69F00",    # Baseline 1
  "fasting" = "#CC79A7",     # Fasting
  "follow-up" = "#0072B2"    # Follow-up
))

color.scale.3 <- scale_fill_manual(values = c(
  "baseline 1" = "#E69F00",    # Baseline 1
  "fasting" = "#CC79A7",     # Fasting
  "follow-up" = "#0072B2"    # Follow-up
))

color.fill.sex <- scale_fill_manual(values = c(
  "male" = "#998ec3",    
  "female" = "#5ab4ac"))
```

## Load and prepare data
```{r}
######################
# Load and prepare phyloseq object
######################

PS.tree <- readRDS("rowdata/PS.tree.rds")

# Changing Geschlecht into Sex
colnames(sample_data(PS.tree))[colnames(sample_data(PS.tree)) == "Geschlecht"] <- "Sex"

# Defining Sex into factors
sample_data(PS.tree)$Sex<- as.factor(sample_data(PS.tree)$Sex)
levels(sample_data(PS.tree)$Sex) <- c("0" = "female", "1" = "male")
sample_data(PS.tree)$Visite<- as.factor(sample_data(PS.tree)$Visite)
levels(sample_data(PS.tree)$Visite) <- c("2" = "baseline 1", "3" = "fasting", "4" = "follow-up")

otu_table_1 <- as.data.frame(otu_table(PS.tree))
sample_table_1 <- as.data.frame(sample_data(PS.tree))
tax_table_1 <- as.data.frame(tax_table(PS.tree))


#saveRDS(PS.tree, file = "rowdata/PS.tree.rds")
```

# Microbial changes to fasting

## Alpha diversity

What are the changes in species diversity due to the fasting intervention?

```{r message=FALSE, warning=FALSE}
# calculate a variaty of diversity indices
tab <- microbiome::alpha(PS.tree, index = "all") 

#####################
## Species richness
####################
tab_richness <- richness(PS.tree)
Samp <- sample_data(PS.tree)
samp_richness <- (merge(Samp, tab_richness, by = 0))

samp_richness %>%
  ggplot(aes(x = Visite, y = chao1, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  color.scale.3 +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  boxplot.theme +
  labs(y = "Chao 1 index") -> fig.2.chao1.fasting
fig.2.chao1.fasting


#### statistics for chao 1 index
samp_richness %>%
  group_by(Visite) %>%
  summarize(mean = mean(chao1),
            sd = sd(chao1))

complete_alpha <- samp_richness %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_alpha %>% 
  pairwise_wilcox_test(
    formula = chao1 ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")


###################
## Alpha diversity
###################

### three timepoints
samp_alpha <- (merge(Samp, tab, by = 0))

samp_alpha %>%
  ggplot(aes(x = Visite, y = diversity_shannon, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  color.scale.3 +
  geom_point(position = position_jitter(width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  labs(y = "Shannon diversity") -> fig.2.shannon.plot
fig.2.shannon.plot


#### statistics for alpha diversity

samp_alpha %>%
  group_by(Visite) %>%
  summarize(mean = mean(diversity_shannon),
            sd = sd(diversity_shannon))


complete_alpha <- samp_alpha %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_alpha %>% 
  pairwise_wilcox_test(
    formula = diversity_shannon ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

#######################
#### Simpson index
####################
samp_alpha %>%
  ggplot(aes(x = Visite, y = diversity_inverse_simpson, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  color.scale.3 +
  geom_point(position = position_jitter(width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  labs(y = "inverse simpson index") -> fig.2.simpson.plot
fig.2.simpson.plot

#### statistics
samp_alpha %>%
  group_by(Visite) %>%
  summarize(mean = mean(diversity_inverse_simpson),
            sd = sd(diversity_inverse_simpson))

complete_alpha <- samp_alpha %>% 
  group_by(Probanden_ID) %>% 
  filter(n() == 3) %>% 
  ungroup()

complete_alpha %>% 
  pairwise_wilcox_test(
    formula = diversity_inverse_simpson ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")

```

## Dependencies of changes in alpha diversity from baseline alpha diversity
```{r message=FALSE, warning=FALSE}
###################
### Prepare data
#################

# separate baseline alpha diversity at baseline and fasting
samp_alpha_baseline <- samp_alpha %>%
  filter(Visite == "baseline 1") %>%
  select(Probanden_ID, diversity_shannon, Sex) %>%
  rename(diversity_shannon_baseline = diversity_shannon)

samp_alpha_fasting <- samp_alpha %>%
  filter(Visite == "fasting") %>%
  select(Probanden_ID, diversity_shannon, Sex, BMI) %>%
  rename(diversity_shannon_fasting = diversity_shannon)

samp_alpha_joined <- samp_alpha_baseline %>%
  full_join(samp_alpha_fasting, by = "Probanden_ID") %>%
  rename(Sex = Sex.x) %>%
  select(-Sex.y)

# create da datafroma which contains the differences in alpha diversity from baseline to fasting (change_fasting)
samp_alpha_joined <- samp_alpha_joined %>%
  mutate(change_fasting = diversity_shannon_fasting - diversity_shannon_baseline)


################################################################################
###### create model to check influence of features on change in alpha diversity
################################################################################
lm(change_fasting ~ diversity_shannon_baseline + Sex +BMI, data = samp_alpha_joined) -> alpha.baseline.sex.bmi
summary(alpha.baseline.sex.bmi) # this result suggest that baseline diversity significantly influences the change in fasting diversity while the effect of sex is not significant after accounting for baseline diversity

## check weather baseline has an influence
lm(change_fasting ~ diversity_shannon_baseline, data = samp_alpha_joined) -> alpha.baseline
summary(alpha.baseline) # this result suggest that baseline diversity significantly influences the change in fasting diversity

## check weather sex has an influence
lm(change_fasting ~ Sex, data = samp_alpha_joined) -> alpha.sex
summary(alpha.sex) #The change in fasting diversity is not significantly influenced by sex alone after adjusting for baseline diversity.

# check weather BMI has an inflience
lm(change_fasting ~ diversity_shannon_baseline + BMI, data = samp_alpha_joined) -> alpha.baseline.BMI
summary(alpha.baseline.BMI)

#### summary of the statistics: changes uppon fasting ####
model.1.fasting <- tidy(alpha.baseline.sex.bmi)
model.2.fasting <- tidy(alpha.baseline)
model.3.fasting <- tidy(alpha.sex)

rbind(model.1.fasting, model.2.fasting, model.3.fasting) -> summary.model.fasting
summary.model.fasting
#In summary, while baseline diversity significantly affects the change in fasting diversity, the effect of sex is not significant once baseline diversity is taken into account. Therefore, the primary factor driving the change in fasting diversity appears to be the initial baseline diversity level.

##############################
####### Plot linear correlation
###############################

##### between shannon baseline alpha diversity and change of shannon diversity
samp_alpha_joined %>%
  ggplot(aes(x = diversity_shannon_baseline, y = change_fasting)) +
  geom_jitter(size = 3, alpha = 0.7) +  # Increase point size and add transparency
  geom_smooth(method = "lm", color = "cornflowerblue") +
  theme_classic() +
  labs(y = "Δ baseline - fasting", x = "baseline") +
  ylim(-1, 2.5) +
  xlim(1.5, 4.75) +
  stat_cor(label.x = 2.8, label.y = 2, size = 7) +  # Add R² value to the plot
  theme(axis.text = element_text(size = 24, color = "black"),  # Adjust axis label size
        axis.title = element_text(size = 24),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.line = element_blank()) -> figure3.lm.fasting.plot
figure3.lm.fasting.plot

cor(samp_alpha_joined$diversity_shannon_baseline, samp_alpha_joined$change_fasting)

####### add overall changes in alpha diversity
samp_alpha_recovery <- samp_alpha %>%
  filter(Visite == "follow-up") %>%
  select(Probanden_ID, diversity_shannon, Sex, BMI) %>%
  rename(diversity_shannon_recovery = diversity_shannon)

samp_alpha_joined <- samp_alpha_joined %>%
  full_join(samp_alpha_recovery, by = "Probanden_ID") %>%
  rename(Sex = Sex.x) %>%
  select(-Sex.y)

samp_alpha_joined <- samp_alpha_joined %>%
  mutate(change_recovery = diversity_shannon_recovery - diversity_shannon_fasting) %>%
  mutate(change_overall = diversity_shannon_recovery -diversity_shannon_baseline)

## plot correlation of overall changes in alpha diversity and shannon baseline diversity
lm(change_overall ~ diversity_shannon_baseline, data = samp_alpha_joined) -> lm.overall
summary(lm.overall)

samp_alpha_joined %>%
  ggplot(aes(x = diversity_shannon_baseline, y = change_overall)) +
  geom_jitter(size = 3, alpha = 0.7) +  # Increase point size and add transparency
  geom_smooth(method = "lm", color = "cornflowerblue") +
  theme_classic() +
  labs(y = "Δ baseline - follow-up", x = "baseline") +
  ylim(-1, 2.5) +
  xlim(1.5, 4.75) +
  stat_cor(label.x = 2.8, label.y = 2, size = 7) +  # Add R² value to the plot
  theme(axis.text = element_text(size = 24, color = "black"),  # Adjust axis label size
        axis.title = element_text(size = 24),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.line = element_blank()) -> figure3.lm.alpha.overall.plot
figure3.lm.alpha.overall.plot

```

## Changes in composition

### Beta diversity
```{r}
#################
###### Plot
################

PS.tree <- readRDS("rowdata/PS.tree.rds")
dist_matrix <- distance(PS.tree, method = "bray")
pcoa <- cmdscale(dist_matrix)
Samp <- data.frame(sample_data(PS.tree))
Samp <- merge(Samp, pcoa, by = 0)
row.names(Samp) <- Samp$Row.names
Samp$Row.names <- NULL

centroids <- aggregate(cbind(V1, V2) ~ Visite, Samp, mean)

fig.2.bray.curtis <- ggplot(Samp, aes(x = V1, y = V2)) +
  stat_ellipse(
    aes(color = Visite, fill = Visite),
    geom = "polygon",
    alpha = 0.1,
    level = 0.95,
    linetype = 1
  ) +
  geom_point(aes(color = Visite), size = 3, alpha = 0.7) +
  geom_point(
    data = centroids,
    aes(fill = Visite),
    size = 7,         # Bigger centroids
    shape = 21,       # Shape that supports both fill and outline
    color = "black"   # Outline
  ) +
  scale_color_manual("Visite",
    values = c("baseline 1" = "#E69F00", 
               "fasting"    = "#CC79A7", 
               "follow-up"  = "#0072B2")
  ) +
  scale_fill_manual("Visite",
    values = c("baseline 1" = "#E69F00", 
               "fasting"    = "#CC79A7", 
               "follow-up"  = "#0072B2")
  ) +
  xlab("Bray-Curtis distance V1") +
  ylab("Bray-Curtis distance V2") +
  theme_classic() +
    theme(
        axis.text.x = element_text(size = 24, color = "black"),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "top")

fig.2.bray.curtis

######################
###### statistics ####
#####################

# Calculating bray_curtis_dist
bray_curtis_dist <- distance(PS.tree, method = "bray")

#per_bray <- adonis(bray_curtis_dist ~ Visite + Sex + Probanden_ID + koerpergewicht + BMI, data = Samp,  strata = Samp$Probanden_ID)
adonis2(
  bray_curtis_dist ~ Visite + Sex + koerpergewicht + BMI,
  data = Samp,
  strata = Samp$Probanden_ID,
  by = "margin"
)

### test different interactions
##############################
## baseline and fasting only
#############################
PS.tree_fasting <- subset_samples(PS.tree, Visite != "follow-up")
bray_curtis_fasting <- distance(PS.tree_fasting, method = "bray")
Samp_fasting <- data.frame(sample_data(PS.tree)) %>%
  filter(Visite != "follow-up")

perm_fasting <- adonis(bray_curtis_fasting ~ Visite * Sex, data = Samp_fasting,  strata = Samp_fasting$Probanden_ID) # the interaction is not significant
perm_fasting$aov.tab
perm_fasting <- adonis(bray_curtis_fasting ~ Visite + Sex, data = Samp_fasting,  strata = Samp_fasting$Probanden_ID) # the interaction is not significant
perm_fasting$aov.tab
perm_fasting <- adonis(bray_curtis_fasting ~ Visite, data = Samp_fasting,  strata = Samp_fasting$Probanden_ID) 
adonis2(bray_curtis_fasting ~ Visite, data = Samp_fasting,  strata = Samp_fasting$Probanden_ID) # the changes from baseline to fasting are significant
perm_fasting$aov.tab
perm_fasting <- adonis(bray_curtis_fasting ~ Sex, data = Samp_fasting,  strata = Samp_fasting$Probanden_ID)
perm_fasting$aov.tab

##############################
#### fasting and recovery only
##############################
PS.tree_recovery <- subset_samples(PS.tree, Visite != "baseline 1")
bray_curtis_recovery <- distance(PS.tree_recovery, method = "bray")

Samp_recovery <- data.frame(sample_data(PS.tree)) %>%
  filter(Visite != "baseline 1")

perm_recovery <- adonis(bray_curtis_recovery ~ Visite * Sex, data = Samp_recovery,  strata = Samp_recovery$Probanden_ID) # the interaction is not significant
perm_recovery$aov.tab
perm_recovery <- adonis(bray_curtis_recovery ~ Visite + Sex, data = Samp_recovery,  strata = Samp_recovery$Probanden_ID) # the interaction is not significant
perm_recovery$aov.tab

perm_recovery <- adonis(bray_curtis_recovery ~ Visite + BMI, data = Samp_recovery,  strata = Samp_recovery$Probanden_ID) # the interaction is not significant
perm_recovery$aov.tab

perm_recovery <- adonis(bray_curtis_recovery ~ Visite, data = Samp_recovery,  strata = Samp_recovery$Probanden_ID)
adonis2(bray_curtis_recovery ~ Visite, data = Samp_recovery,  strata = Samp_recovery$Probanden_ID)# the changes from fasting to follow-up are significant
perm_recovery$aov.tab

##############################
#### fasting and follow-up only
##############################
PS.tree_overall <- subset_samples(PS.tree, Visite != "fasting")
bray_curtis_overall <- distance(PS.tree_recovery, method = "bray")

Samp_overall<- data.frame(sample_data(PS.tree_overall)) %>%
  filter(Visite != "fasting")

perm_overall <- adonis(bray_curtis_overall ~ Visite, data = Samp_overall,  strata = Samp_overall$Probanden_ID) # the changes from fasting to follow-up are significant
adonis2(bray_curtis_overall ~ Visite, data = Samp_overall,  strata = Samp_overall$Probanden_ID)
perm_overall$aov.tab
```


### Trajectorie plots
```{r}
Samp <- data.frame(meta(PS.tree))
bray <- distance(PS.tree, method = "bray")
pcoa <- cmdscale(bray)
Samp <- merge(Samp, pcoa, by = 0)

time_a <- Samp[Samp$Visite == "baseline 1", ]
time_b <- Samp[Samp$Visite == "fasting", ]
time_c <- Samp[Samp$Visite == "follow-up", ]

# join df zu einem df 
all <- merge(time_a, time_b, by = "Probanden_ID", all = TRUE, suffixes = c(".a",".b"))


center <- all
for (i in c("V1.a","V1.b")){
  center[,i] <- (all[, i] - all[, "V1.a"])
}
for (i in c("V2.a","V2.b")){
  center[,i] <- (all[, i] - all[, "V2.a"])
}
center <- center[, c("V1.a","V1.b","V2.a","V2.b", "Probanden_ID")]
center <- na.exclude(center)


arrowCentroids <- as.data.frame(matrix(c(mean(center$V1.b),
                                         mean(center$V2.b), 
                                         center$V1.a[1], center$V2.a[1]), ncol = 4))

# plot centered 
plot_all <- ggplot(center, aes(x = V1.a, y = V2.a)) +
  geom_point(data = center, aes(x = V1.a, y = V2.a), size = 4, alpha = 0.8) + 
  theme_classic() +
  geom_point(data = center, aes(x = V1.b, y = V2.b), color = "#CC79A7", size = 3, alpha = 0.3) +
  # geom_point(data = center, aes(x = V1.c, y = V2.c), color = "#3C5488FF", size = 3, alpha = 0.3) +
  geom_segment(data = center,color ="#CC79A7",
               aes(x = V1.a,
                   y = V2.a,
                   xend = V1.b,
                   yend = V2.b,
                   color = "a to b", alpha = .5),
               arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2), color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), color = "#E69F00", size = 5) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2), color = "#CC79A7" , size = 5) +
  geom_segment(data = arrowCentroids,
               aes(x = V3,
                   y = V4,
                   xend = V1,
                   yend = V2,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black" ) +
  xlab("Bray-Curtis distance V1")+
  ylab("Bray-Curtis distance V2")+
  guides(color = guide_legend(title = "trajectory")) +
  theme (axis.title.x = element_text (size = 24), 
         axis.text.x = element_text (size = 24), 
         axis.text.y = element_text (size = 24), 
         axis.title.y = element_text (size = 24))

print(plot_all)

# fasting to follow-up
time_b <- Samp[Samp$Visite == "fasting", ]
time_c <- Samp[Samp$Visite == "follow-up", ]

all <- merge(time_b, time_c, by = "Probanden_ID", all = TRUE, suffixes = c(".b",".c"))

center <- all
for (i in c("V1.b","V1.c")){
  center[,i] <- (all[, i] - all[, "V1.b"])
}
for (i in c("V2.b","V2.c")){
  center[,i] <- (all[, i] - all[, "V2.b"])
}
center <- center[, c("V1.b","V1.c","V2.b","V2.c", "Probanden_ID")]

center <- na.exclude(center)

arrowCentroids <- as.data.frame(matrix(c(mean(center$V1.c),
                                         mean(center$V2.c), 
                                         center$V1.b[1], center$V2.b[1]), ncol = 4))

# plot centered 
plot_all1 <- ggplot(center, aes(x = V1.b, y = V2.b)) +
  geom_point(data = center, aes(x = V1.b, y = V2.b), size = 4, alpha = 0.8) + 
  theme_classic() +
  geom_point(data = center, aes(x = V1.c, y = V2.c), color = "#0072B2", size = 3, alpha = 0.3) +
  # geom_point(data = center, aes(x = V1.c, y = V2.c), color = "#3C5488FF", size = 3, alpha = 0.3) +
  geom_segment(data = center,color = "#0072B2",
               aes(x = V1.b,
                   y = V2.b,
                   xend = V1.c,
                   yend = V2.c,
                   color = "b to c", alpha = .5),
               arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2), color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), color = "#E69F00", size = 5) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2), color = "#0072B2" , size = 5) +
  geom_segment(data = arrowCentroids,
               aes(x = V3,
                   y = V4,
                   xend = V1,
                   yend = V2,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black" ) +
  xlab("Bray-Curtis distance V1")+
  ylab("Bray-Curtis distance V2")+
  guides(color = guide_legend(title = "trajectory")) +
  # scale_color_manual(values = c("#4DBBD5FF", "#00A087FF", "black"))+
  
  theme (axis.title.x = element_text (size = 24), 
         axis.text.x = element_text (size = 24), 
         axis.text.y = element_text (size = 24), 
         axis.title.y = element_text (size = 24))

print(plot_all1)

time_b <- Samp[Samp$Visite == "fasting", ]
time_c <- Samp[Samp$Visite == "follow-up", ]


all.1 <- merge(time_b, time_c, by = "Probanden_ID", all = TRUE, suffixes = c(".b",".c"))

center.1 <- all.1
for (i in c("V1.b","V1.c")){
  center.1[,i] <- (all.1[, i] - all.1[, "V1.b"])
}
for (i in c("V2.b","V2.c")){
  center.1[,i] <- (all.1[, i] - all.1[, "V2.b"])
}
center.1 <- center.1[, c("V1.b","V1.c","V2.b","V2.c", "Probanden_ID")]

center.1 <- na.exclude(center.1) # nur Donoren die alle swabsite haben

arrowCentroids.1 <- as.data.frame(matrix(c(mean(center.1$V1.c),
                                           mean(center.1$V2.c), 
                                           center.1$V1.b[1], center.1$V2.b[1]), ncol = 4))




plot_all +
  geom_point(data = center.1, aes(x = V1.b, y = V2.b), size = 4, alpha = 0.8, color = "#E69F00") + 
  theme_classic() +
  geom_point(data = center.1, aes(x = V1.c, y = V2.c), color = "#0072B2", size = 3, alpha = 0.3) +
  # geom_point(data = center, aes(x = V1.c, y = V2.c), color = "#3C5488FF", size = 3, alpha = 0.3) +
  geom_segment(data = center.1,color = "#0072B2",
               aes(x = V1.b,
                   y = V2.b,
                   xend = V1.c,
                   yend = V2.c,
                   color = "b to c", alpha = .5),
               arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids.1, aes(x = V1, y = V2), color = "black", size = 6) +
  geom_point(data = arrowCentroids.1, aes(x = V3, y = V4), color = "black", size = 6, fill = "black") +
  geom_point(data = arrowCentroids.1, aes(x = V3, y = V4), color = "black", size = 5) +
  geom_point(data = arrowCentroids.1, aes(x = V1, y = V2), color = "#0072B2" , size = 5) +
  geom_segment(data = arrowCentroids.1,
               aes(x = V3,
                   y = V4,
                   xend = V1,
                   yend = V2,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black" ) +
  xlab("Bray-Curtis distance V1")+
  ylab("Bray-Curtis distance V2")+
  guides(color = guide_legend(title = "trajectory")) +
  boxplot.theme.2 +
  theme(legend.position = "top")+
  xlab("Bray-Curtis distance V1") -> fig.2.trajectorie.plot

fig.2.trajectorie.plot
```

```{r}
time_a <- Samp[Samp$Visite == "baseline 1", ]
time_b <- Samp[Samp$Visite == "fasting", ]
time_c <- Samp[Samp$Visite == "follow-up", ]


all_three <- merge(time_a, time_b, by = "Probanden_ID", 
                   suffixes = c(".a",".b"))

all_three <- merge(all_three, time_c, by = "Probanden_ID", 
                   suffixes = c("", ".c"))


######################
# Center on baseline
####################

# Temporarily store baseline coordinates
b0_x <- all_three$V1.a
b0_y <- all_three$V2.a

# Subtract baseline from baseline, fasting, follow-up
all_three$V1.a <- all_three$V1.a - b0_x   
all_three$V2.a <- all_three$V2.a - b0_y

all_three$V1.b <- all_three$V1.b - b0_x   # fasting is relative to baseline
all_three$V2.b <- all_three$V2.b - b0_y

all_three$V1   <- all_three$V1   - b0_x   # follow-up is relative to baseline
all_three$V2   <- all_three$V2   - b0_y

centroids <- data.frame(
  group = c("baseline", "fasting", "follow-up"),
  x = c(0, mean(all_three$V1.b, na.rm = TRUE), mean(all_three$V1, na.rm = TRUE)),
  y = c(0, mean(all_three$V2.b, na.rm = TRUE), mean(all_three$V2, na.rm = TRUE))
)


ggplot(all_three) +
  geom_segment(
    aes(x = V1.a, y = V2.a, xend = V1.b, yend = V2.b),
    color = "#CC79A7",
    arrow = arrow(length = unit(0.3, "cm")),
    alpha = 0.7
  ) +
  geom_segment(
    aes(x = V1, y = V2, xend = V1.a, yend = V2.a),
    color = "#0072B2",
    arrow = arrow(length = unit(0.3, "cm")),
    alpha = 0.7
  ) +
  geom_point(aes(x = V1.a, y = V2.a), color = "#E69F00", alpha = 0.7, size = 3) +
  geom_point(aes(x = V1.b, y = V2.b), color = "#CC79A7", size = 3) +
  geom_point(aes(x = V1, y = V2), color = "#0072B2", size = 3) +
  geom_point(
    data = centroids,
    aes(x = x, y = y, fill = group),
    shape = 21,
    size = 6,
    color = "black",
    stroke = 1.5
  ) +
  scale_fill_manual(
    values = c("baseline" = "#E69F00", "fasting" = "#CC79A7", "follow-up" = "#0072B2")
  ) +
  theme_classic() +
  xlab("Bray-Curtis distance V1") +
  ylab("Bray-Curtis distance V2") +
  theme(
    axis.text  = element_text(size = 16, color = "black"),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.position = "top"
  ) -> fig.2.trajectorie.plot.new
fig.2.trajectorie.plot.new
```


## Changes in bacterial abundance

### Significant changes on species level
```{r}
genus <- tax_glom(physeq = PS.tree, taxrank = "Species")
data_biom_all <- as.data.frame(tax_table(PS.tree)) 
data_biom_tax <- as.data.frame(tax_table(genus)) 
data_biom <-as.data.frame(otu_table(genus)) 


data_biom_tax$hash <- rownames(data_biom_tax)
data_biom$hash <- rownames(data_biom)
joined_genus <- left_join(data_biom_tax, data_biom)
joined_genus$hash <- NULL


genus <- joined_genus[, - c(1:6,8:9) ]
rownames(genus) <- make.names(genus$Species, unique = T)
genus$Species <- NULL
genus <- as.data.frame(t(genus))


sample.data <- data.frame(sample_data(PS.tree)) %>%
  select(Probanden_ID, Visite)

## combine with bacterial data
names <- readRDS("rowdata/names.rds")
genus <- t(names)
bacteria.all <- cbind(sample.data, genus)

## make long format
bacteria.long <- pivot_longer(bacteria.all, cols = c(3:ncol(bacteria.all)), names_to = "Species", values_to = "Abundance")

# Filter data to only include subjects with all three visits for each species
df_complete <- bacteria.long %>%
  group_by(Species, Probanden_ID) %>%
  filter(n_distinct(Visite) == 3) %>%  # Keep only subjects with all 3 visits
  ungroup()

# Perform paired pairwise Wilcoxon tests across all visits
result <- df_complete %>% 
  group_by(Species) %>%
  pairwise_wilcox_test(
    formula = Abundance ~ Visite, 
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "BH"
  )

# Extract significant results (species that are significant at least once)
signif_results <- result %>%
  filter(p.adj < 0.1)

print(signif_results)
sig_species <- unique(signif_results$Species)

filtered_result <- result %>%
  filter(Species %in% sig_species)

# Compute Cliff's delta for each species and each comparison
all_comps_for_sig_species <- filtered_result %>%
  rowwise() %>%
  mutate(
    cliff_delta = effsize::cliff.delta(
      bacteria.long$Abundance[bacteria.long$Species == Species & bacteria.long$Visite == group2],
      bacteria.long$Abundance[bacteria.long$Species == Species & bacteria.long$Visite == group1],
    )$estimate
  ) %>%
  ungroup() %>%
  mutate(
    is_significant = p.adj < 0.1,
    Direction = case_when(
      cliff_delta > 0 ~ "Increased",
      cliff_delta < 0 ~ "Decreased",
      TRUE            ~ "No change"
    ),
    effect_size = cliff_delta,
    point_alpha = if_else(is_significant, 1, 0.4),
    outline_color = if_else(is_significant, "black", "grey"),
    Comparison = paste(group1, group2, sep = " vs ")
  )

# Clean up species names
sig_plot_data <- all_comps_for_sig_species %>%
  mutate(
    Species = gsub("\\.sp\\.", "##SP##", Species),
    Species = gsub("\\.", " ", Species),
    Species = gsub("##SP##", " sp.", Species),
    Species = gsub("[\\(\\)]", "", Species),
    Species = gsub("\\s+", " ", Species),
    Species = sub("^NA\\s+", "", Species),
    Species = trimws(Species)
  )

sig_plot_data$Comparison <- factor(sig_plot_data$Comparison,
  levels = c("baseline 1 vs fasting", "fasting vs follow-up", "baseline 1 vs follow-up"))

# Save as table for supplement
library(writexl)
#write_xlsx(sig_plot_data, "sig_plot_data_species.xlsx")

order_species <- sig_plot_data %>%
  filter(Comparison == "baseline 1 vs fasting") %>%
  arrange(effect_size) %>%
  pull(Species)

sig_plot_data <- sig_plot_data %>%
  mutate(Species = factor(Species, levels = order_species))



# Plot the results using ggplot2
ggplot(sig_plot_data, aes(
  x = Comparison,
  y = Species  
)) +
  geom_point(
    aes(
      fill = Direction,         
      size = abs(effect_size),  
      alpha = point_alpha,      
      color = outline_color     
    ),
    shape = 21,    
    stroke = 0.8
  ) +
  scale_fill_manual(values = c("Increased" = "#0075DC", 
                               "Decreased" = "#E64B35FF",
                               "No change"  = "grey70")) +
  scale_color_identity() +
  scale_alpha_identity() +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    x = "Comparison",
    y = "Species",
    fill = "Direction",
    size = "|\u0394 Cliff's delta|"
  ) +
  theme_classic(base_size = 24) +
  theme(
    axis.text.y = element_text(face = "italic", color = "black"),
    axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
    axis.title  = element_text(face = "bold"),
    axis.title.x = element_blank(),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) -> fig.2.sig.species.plot

print(fig.2.sig.species.plot)

ggsave(plot = fig.2.sig.species.plot, filename = "fig.2.sig.species.plot.pdf", width = 12.5, height = 10)


```




### Significant changes presented as boxplots
```{r}
# filter the significant species names
significant <- signif_results %>%
  filter(group1 == "baseline 1" & group2 == "follow-up") 

sign.species <- significant$Species

t.names <- data.frame(t(names), check.names = FALSE)
sign.names <- t.names %>%
  select(sign.species)

# bind it with metadata 
sample.data <- data.frame(sample_data(PS.tree)) %>%
  select(Probanden_ID, Visite)

sig.species.meta <- cbind(sample.data, sign.names)

sig.species.long <- pivot_longer(sig.species.meta, cols = c(3:9), names_to = "Species", values_to = "Abundance")

sig.species.long<- sig.species.long %>%
  mutate(
    # Temporarily protect ".sp." by replacing with a placeholder
    Species = gsub("\\.sp\\.", "##SP##", Species),
    # Replace all remaining dots with a space
    Species = gsub("\\.", " ", Species),
    # Restore the placeholder back to " sp."
    Species = gsub("##SP##", " sp.", Species),
    # Remove any parentheses (both opening and closing)
    Species = gsub("[\\(\\)]", "", Species),
    # Clean up multiple spaces and trim any leading/trailing whitespace
    Species = gsub("\\s+", " ", Species),
    Species = trimws(Species)
  )


## create boxplots 
sig.species.long %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  theme_classic() +
  facet_wrap(~ Species, ncol = 3) +
  ylim(c(0,5))+
  theme(axis.text = element_text(size=20, color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, size=20, color = "black"),
        axis.title = element_text(size=20, color = "black"),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20), 
        axis.title.x = element_blank(),
        panel.border = element_rect(color = "black", fill = NA,  size = 1),
        strip.background = element_blank(), 
        strip.text = element_text(size = 18, color = "black", face = "italic"),
        legend.position = "none") +
  labs(y = "log (Abundance)") -> fig.2.species.significant.boxplots
fig.2.species.significant.boxplots


### everything seperate
sig.species.long %>%
  filter(Species == "Adlercreutzia equolifaciens") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Adlercreutzia equolifaciens", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Adlercreutzia.equolifaciens

ggsave("fig.2.boxplots.Adlercreutzia.equolifaciens.pdf", plot = fig.2.boxplots.Adlercreutzia.equolifaciens, width = 5, height = 5)

sig.species.long %>%
  filter(Species == "Bifidobacterium bifidum CAG 234") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Bifidobacterium bifidum", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Bifidobacterium.bifidum

ggsave("fig.2.boxplots.Bifidobacterium.bifidum.pdf", plot = fig.2.boxplots.Bifidobacterium.bifidum, width = 5, height = 5)

sig.species.long %>%
  filter(Species == "Erysipelatoclostridium ramosum") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Erysipelatoclostridium ramosum", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Erysipelatoclostridium.ramosum

ggsave("fig.2.boxplots.Erysipelatoclostridium.ramosum.pdf", plot = fig.2.boxplots.Erysipelatoclostridium.ramosum, width = 5, height = 5)

sig.species.long %>%
  filter(Species == "NA Acinetobacter sp.CAG 196 CAG 196_36_41") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Acinetobacter sp.CAG 196", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Acinetobacter.CAG

ggsave("fig.2.boxplots.Acinetobacter.CAG.pdf", plot = fig.2.boxplots.Acinetobacter.CAG, width = 5, height = 5)

sig.species.long %>%
  filter(Species == "NA Azospirillum species incertae sedis") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Azospirillum sp. incertae sedis", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Azospirillum.sp

ggsave("fig.2.boxplots.Azospirillum.sp.pdf", plot = fig.2.boxplots.Azospirillum.sp, width = 5, height = 5)

sig.species.long %>%
  filter(Species == "NA Butyricicoccus sp.UC5 1 1D1") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Butyricicoccus sp.UC5", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Butyricicoccus.sp.

ggsave("fig.2.boxplots.Butyricicoccus.sp.pdf", plot = fig.2.boxplots.Butyricicoccus.sp., width = 5, height = 5)

sig.species.long %>%
  filter(Species == "Ruthenibacterium lactatiformans 4_3_54A2FAA") %>%
  ggplot(aes(x = Visite, y = log(Abundance), fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Ruthenibacterium lactatiformans", y = "log (Abundance)") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Ruthenibacterium.lactatiformans

ggsave("fig.2.boxplots.Ruthenibacterium.lactatiformans.pdf", plot = fig.2.boxplots.Ruthenibacterium.lactatiformans, width = 5, height = 5)
```



### Significant changes on genus level

```{r}
genus <- tax_glom(physeq = PS.tree, taxrank = "Genus")
data_biom_all <- as.data.frame(tax_table(PS.tree)) 
data_biom_tax <- as.data.frame(tax_table(genus)) 
data_biom <-as.data.frame(otu_table(genus)) 


#genus_otu_short <- as.data.frame(otu_table(data_biom_tax))
data_biom_tax$hash <- rownames(data_biom_tax)
data_biom$hash <- rownames(data_biom)
joined_genus <- left_join(data_biom_tax, data_biom)
joined_genus$hash <- NULL

# Extract genus and abundance
genus <- joined_genus[, - c(1:5,7:8) ]
rownames(genus) <- make.names(genus$Genus, unique = T)
genus$Genus <- NULL
genus <- as.data.frame(t(genus))

## add relevant meta data
sample.data <- data.frame(sample_data(PS.tree)) %>%
  select(Probanden_ID, Visite)

## combine with bacterial data
bacteria.all <- cbind(sample.data, genus)

## make long format
bacteria.long <- pivot_longer(bacteria.all, cols = c(3:ncol(bacteria.all)), names_to = "Genus", values_to = "Abundance")

# Filter data to only include subjects with all three visits for each species
df_complete <- bacteria.long %>%
  group_by(Genus, Probanden_ID) %>%
  filter(n_distinct(Visite) == 3) %>%  # Keep only subjects with all 3 visits
  ungroup()

# Now perform the paired pairwise Wilcoxon tests across all visits
result.genus <- df_complete %>% 
  group_by(Genus) %>%
  pairwise_wilcox_test(
    formula = Abundance ~ Visite, 
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "BH"
  )

### extract signifcant results
signif.results.genus <- result.genus %>%
  filter(p.adj < 0.1)

print(signif_results)

# Species that are significant at least once
sig_genus <- unique(signif.results.genus$Genus)

filtered_result <- result.genus %>%
  filter(Genus %in% sig_genus)

# Compute Cliff's delta for each Genus and each comparison
all_comps_for_sig_Genus <- filtered_result %>%
  rowwise() %>%
  mutate(
    cliff_delta = effsize::cliff.delta(
      bacteria.long$Abundance[bacteria.long$Genus == Genus & bacteria.long$Visite == group2],
      bacteria.long$Abundance[bacteria.long$Genus == Genus & bacteria.long$Visite == group1]
    )$estimate
  ) %>%
  ungroup() %>%
  mutate(
    is_significant = p.adj < 0.1,
    Direction = case_when(
      cliff_delta > 0 ~ "Increased",
      cliff_delta < 0 ~ "Decreased",
      TRUE            ~ "No change"
    ),
    effect_size = cliff_delta,
    point_alpha = if_else(is_significant, 1, 0.4),
    outline_color = if_else(is_significant, "black", "grey"),
    Comparison = paste(group1, group2, sep = " vs ")
  )

# Clean up Genus names
sig_plot_data <- all_comps_for_sig_Genus %>%
  mutate(
    # Temporarily protect ".sp." by replacing with a placeholder
    Genus = gsub("\\.sp\\.", "##SP##", Genus),
    # Replace all remaining dots with a space
    Genus = gsub("\\.", " ", Genus),
    # Restore the placeholder back to " sp."
    Genus = gsub("##SP##", " sp.", Genus),
    # Remove any parentheses
    Genus = gsub("[\\(\\)]", "", Genus),
    # Clean up extra spaces and trim
    Genus = gsub("\\s+", " ", Genus),
    Genus = trimws(Genus)
  )

# Order comparisons on x-axis 
sig_plot_data$Comparison <- factor(sig_plot_data$Comparison,
  levels = c("baseline 1 vs fasting", "fasting vs follow-up", "baseline 1 vs follow-up"))

# Save as table for supplement
#write_xlsx(sig_plot_data, "sig_plot_data_Genus.xlsx")

# Extract ordering for Genus based on the effect size from "baseline 1 vs fasting"
order_Genus <- sig_plot_data %>%
  filter(Comparison == "baseline 1 vs fasting") %>%
  arrange(effect_size) %>%
  pull(Genus)

# Update the Genus factor levels based on the ordering from the baseline 1 vs fasting comparison
sig_plot_data <- sig_plot_data %>%
  mutate(Genus = factor(Genus, levels = order_Genus))

# Plot the results
ggplot(sig_plot_data, aes(
  x = Comparison,
  y = Genus  # Now already ordered based on baseline 1 vs fasting
)) +
  geom_point(
    aes(
      fill = Direction,         
      size = abs(effect_size),  
      alpha = point_alpha,      
      color = outline_color     
    ),
    shape = 21,    
    stroke = 0.8
  ) +
  scale_fill_manual(values = c("Increased" = "#0075DC", 
                               "Decreased" = "#E64B35FF",
                               "No change"  = "grey70")) +
  scale_color_identity() +
  scale_alpha_identity() +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    x = "Comparison",
    y = "Genus",
    fill = "Direction",
    size = "|\u0394 Cliff's delta|"
  ) +
  theme_classic(base_size = 24) +
  theme(
    axis.text.y = element_text(face = "italic", color = "black"),
    axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
    axis.title  = element_text(face = "bold"),
    axis.title.x = element_blank(),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) -> fig.2.sig.Genus.plot

print(fig.2.sig.Genus.plot)
#ggsave(plot = fig.2.sig.Genus.plot, filename = "fig.2.sig.Genus.plot.pdf", width = 14, height = 18)
```

## Changes in the functional potential

### Preparing the data
```{r}
# Load the kegg names
func <- read.table("rowdata/functions/kegg_names_annotation.tsv", sep = "\t", header =T)

# replace the symbols which are not "good" for R
func$Long_names <- gsub(" ", "_", func$Long_names)
func$Long_names <- gsub(" ", "_", func$Long_names)
func$Long_names <- gsub(" ", "_", func$Long_names)
func$Long_names <- gsub("+", "", func$Long_names)
func$Long_names <- gsub("-", "_", func$Long_names)
func$Long_names <- gsub("/", "_", func$Long_names)

nice_feature_names_for_annotation <- func %>%
  select(c("KO_number", "Long_names"))
# write.table(nice_feature_names_for_annotation, file = "~/Dropbox/Master_arbeit_Mina/healthy_fast/functions/helper_function_names.tsv", sep="\t", quote=F)

func$Long_names<- NULL
row.names(func) <-func$KO_number
func$KO_number <- NULL

func <- as.data.frame(t(func))
func$rowname <- row.names(func)
row_names_new <- gsub(pattern = "preprocessed\\.|filtered_qc\\.pair", replacement = "", rownames(func))
row_names_new <- gsub(pattern = "_", replacement = ".", row_names_new)
rownames(func) <- row_names_new


# Check if each column is numeric
numeric_cols <- sapply(func, is.numeric)

# Find columns that are not numeric
non_numeric_cols <- names(numeric_cols)[!numeric_cols]

# Print non-numeric column names
print(non_numeric_cols)

## the output was that rowname is non numeric, therefore we have to delete that colum

func$rowname <- NULL

# the dataset still contains V1 samples which need to be filtered out
fun <- func %>%
  rownames_to_column(var = "row_names")

fun <- fun %>%
  filter(!grepl("^V1", row_names))

fun <- fun %>%
  column_to_rownames(var = "row_names")

fun$rowname <- NULL

#write.table(fun, file = "rowdata/functions/functions_rdy.tsv", sep="\t", quote=F)
```

### Combine functional data with meta data
```{r}
#Loading preprocessed data
functions_table <- read.table("rowdata/functions/functions_rdy.tsv", sep="\t", header = TRUE)

# make sure that we have the same patients included in the functions table
sample_table_1 <- data.frame(sample_data(PS.tree))
functions_table <- functions_table[rownames(functions_table) %in% rownames(sample_table_1), ] 

functions_table_names <- t(functions_table)
functions_table_names <- as.data.frame(functions_table_names)
functions_table_names <- functions_table_names %>%
  rownames_to_column(var = "KO_number")

## nice names: 
nice <- nice_feature_names_for_annotation
nice <- as.data.frame(nice)
row.names(nice) <- NULL

nice <- nice %>%
  mutate(Long_names = coalesce(Long_names, "no_name"))

nice$Long_names <- gsub("_\\[EC:.*?\\]", "", nice$Long_names)


kegg_names <- read_csv("rowdata/functions/kegg_names_annotation_new.csv")
kegg_names <- kegg_names[ ,c(1, ncol(kegg_names))]

kegg_names <- kegg_names %>%
  mutate(Long_names = coalesce(Long_names, "no_name")) %>%
  rename("KO_number" = 1)

kegg_names$KO_number <- gsub("ko:", "", kegg_names$KO_number)


kegg_modules <- read_csv("rowdata/functions/kegg_module_annotation.csv")



functions_table_names <- functions_table_names %>%
  left_join(nice, by = "KO_number")

# summarizing samples with same Long name
functions_table_names_summary <- functions_table_names %>%
  group_by(Long_names) %>%
  summarize(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))


functions_table_names <- functions_table_names_summary %>%
  column_to_rownames(var = "Long_names")

#functions_table_names <- t(functions_table_names)

```

### Alpha diversity of functional potential
```{r}
alpha <- microbiome::alpha(t(functions_table), index = "all") 

## merge with metadata
functions_alpha <- merge(sample_table_1, alpha, by = 0)
functions_alpha <- functions_alpha %>%
  column_to_rownames(var = "Row.names")


functions_alpha %>%
  ggplot(aes(x = Visite, y = diversity_shannon, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  boxplot.theme +
  labs(y = "shannon diversity") -> fig.2.shannon.functions.plot

fig.2.shannon.functions.plot
```
### Beta diversity of functional potential
```{r}
# Calculate Bray-Curtis distance using the vegan package
dist_bc <- as.matrix((vegdist(x = functions_table , method = "bray"))) #here we have to switch rows and columns because we have taxa_are_rows = TRUE
dist_bc[1:5, 1:5]

pcoa <- cmdscale(dist_bc)

Samp <- data.frame(sample_data(PS.tree))
Samp <- merge(Samp, pcoa, by = 0)
row.names(Samp) <- Samp$Row.names
Samp$Row.names <- NULL

centroids <- aggregate(cbind(V1, V2) ~ Visite, Samp, mean)

ggplot(Samp, aes(x = V1, y = V2)) +
  stat_ellipse(
    aes(color = Visite, fill = Visite),
    geom = "polygon",
    alpha = 0.1,
    level = 0.95,
    linetype = 1
  ) +
  geom_point(aes(color = Visite), size = 3, alpha = 0.7) +
  geom_point(
    data = centroids,
    aes(fill = Visite),
    size = 7,         
    shape = 21,       
    color = "black"   
  ) +
  scale_color_manual("Visite",
    values = c("baseline 1" = "#E69F00", 
               "fasting"    = "#CC79A7", 
               "follow-up"  = "#0072B2")
  ) +
  scale_fill_manual("Visite",
    values = c("baseline 1" = "#E69F00", 
               "fasting"    = "#CC79A7", 
               "follow-up"  = "#0072B2")
  ) +
  xlab("Bray-Curtis distance V1") +
  ylab("Bray-Curtis distance V2") +
  boxplot.theme.2 +
  theme(
    axis.title.x = element_text(size = 24, color = "black"),  # restore x-axis title
    axis.line = element_line(color = "black"),
    legend.position = "top") -> fig.2.bray.curtis.functions

fig.2.bray.curtis.functions
```


### Significant changes in functional potential
```{r}
Samp <- data.frame(meta(PS.tree))
samp_for_model <- Samp[, c("Probanden_ID", "koerpergewicht", "Sex", "Visite", "waist", "Alter")]

kegg_modules <- read_csv("rowdata/functions/kegg_module_annotation.csv")

kegg_modules <- kegg_modules[ ,-c(1, 128:133)]

functions_modules_summary <- kegg_modules %>%
  group_by(long) %>%
  summarize(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))

functions_modules_summary <- na.omit(functions_modules_summary)

kegg_modules <- functions_modules_summary  %>%
  column_to_rownames(var = "long")

colnames(kegg_modules) <- gsub(colnames(kegg_modules), pattern = "preprocessed\\.|filtered_qc\\.pair", replacement = "")
colnames(kegg_modules) <- gsub(colnames(kegg_modules), pattern = "_", replacement = ".")

kegg_modules <- data.frame(t(kegg_modules))

## remove the samples which are not present in the sample data
kegg_modules <- kegg_modules[rownames(kegg_modules) %in% rownames(Samp), ]

modules_meta <- merge(samp_for_model, kegg_modules, by = 0)

modules_meta <- modules_meta %>%
  column_to_rownames(var = "Row.names")

#saveRDS(modules_meta, file = "modules.meta.leaner.rds")

functions_long <- pivot_longer(modules_meta, cols = c(7:ncol(modules_meta)), values_to = "Abundance", names_to = "Function")

# Filter data to only include subjects with all three visits for each Genus
df_complete <- functions_long %>%
  group_by(Function, Probanden_ID) %>%
  filter(n_distinct(Visite) == 3) %>%  # Keep only subjects with all 3 visits
  ungroup()

# Now perform the paired pairwise Wilcoxon tests across all visits
result.function <- df_complete %>% 
  group_by(Function) %>%
  pairwise_wilcox_test(
    formula = Abundance ~ Visite, 
    paired = TRUE,
    id = "Probanden_ID",
    p.adjust.method = "BH"
  )

### extract signifcant results
signif.results.function <- result.function %>%
  filter(p.adj < 0.1)

# Genus that are significant at least once
sig_function <- unique(signif.results.function$Function)

filtered_result <- result.function %>%
  filter(Function %in% sig_function)


# Compute Cliff's delta for each Function and each comparison
all_comps_for_sig_function <- filtered_result %>%
  rowwise() %>%
  mutate(
    cliff_delta = effsize::cliff.delta(
      functions_long$Abundance[functions_long$Function == Function & functions_long$Visite == group2],
      functions_long$Abundance[functions_long$Function == Function & functions_long$Visite == group1]
    )$estimate
  ) %>%
  ungroup() %>%
  mutate(
    is_significant = p.adj < 0.1,
    Direction = case_when(
      cliff_delta > 0 ~ "Increased",
      cliff_delta < 0 ~ "Decreased",
      TRUE            ~ "No change"
    ),
    effect_size = cliff_delta,
    point_alpha = if_else(is_significant, 1, 0.4),
    outline_color = if_else(is_significant, "black", "grey"),
    Comparison = paste(group1, group2, sep = " vs ")
  )

# Clean up Function names
sig_plot_data <- all_comps_for_sig_function %>%
  mutate(
    # Temporarily protect ".sp." by replacing with a placeholder
    Function = gsub("\\.sp\\.", "##SP##", Function),
    # Replace all remaining dots with a space
    Function = gsub("\\.", " ", Function),
    # Restore the placeholder back to " sp."
    Function = gsub("##SP##", " sp.", Function),
    # Remove any parentheses
    Function = gsub("[\\(\\)]", "", Function),
    # Clean up extra spaces and trim
    Function = gsub("\\s+", " ", Function),
    Function = trimws(Function)
  )

# Order comparisons on x-axis (optional)
sig_plot_data$Comparison <- factor(sig_plot_data$Comparison,
  levels = c("baseline 1 vs fasting", "fasting vs follow-up", "baseline 1 vs follow-up"))

# Save as table for supplement
#write_xlsx(sig_plot_data, "sig_plot_data_Function.xlsx")

# Extract ordering for Function based on the effect size from "baseline 1 vs fasting"
order_Function <- sig_plot_data %>%
  filter(Comparison == "baseline 1 vs fasting") %>%
  arrange(effect_size) %>%
  pull(Function)

# Update the Function factor levels based on the ordering from the baseline 1 vs fasting comparison
sig_plot_data <- sig_plot_data %>%
  mutate(Function = factor(Function, levels = order_Function))

# Plot the results using ggplot2
ggplot(sig_plot_data, aes(
  x = Comparison,
  y = Function  
)) +
  geom_point(
    aes(
      fill = Direction,        
      size = abs(effect_size),  
      alpha = point_alpha,      
      color = outline_color     
    ),
    shape = 21,    
    stroke = 0.8
  ) +
  scale_fill_manual(values = c("Increased" = "#0075DC", 
                               "Decreased" = "#E64B35FF",
                               "No change"  = "grey70")) +
  scale_color_identity() +
  scale_alpha_identity() +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    x = "Comparison",
    y = "Function",
    fill = "Direction",
    size = "|\u0394 Cliff's delta|"
  ) +
  theme_classic(base_size = 24) +
  theme(
    axis.text.y = element_text(face = "italic", color = "black"),
    axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
    axis.title  = element_text(face = "bold"),
    axis.title.x = element_blank(),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) -> fig.2.sig.Function.plot

print(fig.2.sig.Function.plot)
ggsave(plot = fig.2.sig.Function.plot, filename = "fig.2.sig.Function.plot.pdf", width = 14, height = 16)

### plot the 3 significant changes
sig.overall <- sig_plot_data %>%
  filter(group1 == "baseline 1" & group2 == "follow-up" & p.adj <= 0.1)

functions_long %>%
  filter(Function == "Ketone.body.biosynthesis") %>%
  ggplot(aes(x = Visite, y = Abundance, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  scale_y_continuous(labels = scales::scientific) +
  #ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Ketone body biosynthesis", y = "functional potential") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Ketone.body.biosynthesis

#ggsave("fig.2.boxplots.Ketone.body.biosynthesis.pdf", plot = fig.2.boxplots.Ketone.body.biosynthesis, width = 5, height = 5)

functions_long %>%
  filter(Function == "Osmoprotectant.transport.system") %>%
  ggplot(aes(x = Visite, y = Abundance, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  scale_y_continuous(labels = scales::scientific) +
  #ylim(c(0,5)) +
  theme_classic() +
  labs(title = "Osmoprotectant transport system", y = "functional potential") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.Osmoprotectant.transport.system

#ggsave("fig.2.boxplots.Osmoprotectant.transport.system.pdf", plot = fig.2.boxplots.Osmoprotectant.transport.system, width = 5, height = 5)

functions_long %>%
  filter(Function == "acetylneuraminate.and.acetylmannosamine.degradation") %>%
  ggplot(aes(x = Visite, y = Abundance, fill = Visite)) +
  geom_line(aes(group = Probanden_ID), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c("baseline 1"= "#E69F00", "fasting"="#CC79A7", "follow-up" = "#0072B2")) +
  geom_point(size = 3,position = position_jitter(width = 0), alpha=0.5) +
  scale_y_continuous(labels = scales::scientific) +
  #ylim(c(0,5)) +
  theme_classic() +
  labs(title = "acetylneuraminate and acetylmannosamine degradation", y = "functional potential") +
  theme_classic() +
    theme(
      plot.title = element_text(size = 24, color = "black", face = "italic"),
        axis.text.x = element_text(size = 24, color = "black", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 24, color = "black"),
        axis.title = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 24, color = "black"),
        legend.title = element_text(size = 24, color = "black"),
        axis.line = element_line(size = 1),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") -> fig.2.boxplots.acetylneuraminate.and.acetylmannosamine.degradation

#ggsave("fig.2.boxplots.acetylneuraminate.and.acetylmannosamine.degradation.pdf", plot = fig.2.boxplots.acetylneuraminate.and.acetylmannosamine.degradation, width = 5, height = 5)

```

## Comparison between baseline 0 and baseline 1

### alpha diversity

```{r}
# load physoseq with all visits
PS.all.visits <- readRDS("rowdata/PS_all_Visits.rds")
PS.V1.V2 <- subset_samples(PS.all.visits, Visite %in% c("V1", "V2"))

PS.V1.V2@sam_data$Visite <- factor(PS.V1.V2@sam_data$Visite, 
                                        levels = c("V1", "V2"),
                                        labels = c("baseline 0", "baseline 1"))

tab <- microbiome::alpha(PS.V1.V2, index = "all") 

samp_V1.V2 <- data.frame(sample_data(PS.V1.V2))

samp_alpha_V1.V2 <- (merge(samp_V1.V2, tab, by = 0))

samp_alpha_V1.V2 %>%
  ggplot(aes(x = Visite, y = diversity_shannon, fill = Visite)) +
  geom_line(aes(group = proband_id), color = "black", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c(
  "baseline 0" = "#FFD700",
  "baseline 1" = "#E69F00")) +
  geom_point(position = position_jitter(width = 0), alpha=0.5, size = 3) +
  boxplot.theme +
  labs(y = "Shannon diversity") -> fig.2.alpha.baseline.comparison
fig.2.alpha.baseline.comparison

#ggsave("fig.2.alpha.baseline.comparison.pdf", plot = fig.2.alpha.baseline.comparison, width = 5, height = 5)

#### statistics for alpha diversity

samp_alpha_V1.V2 %>%
  group_by(Visite) %>%
  summarize(mean = mean(diversity_shannon),
            sd = sd(diversity_shannon))


complete_alpha <- samp_alpha_V1.V2 %>% 
  group_by(proband_id) %>% 
  filter(n() == 2) %>% 
  ungroup()

complete_alpha %>% 
  pairwise_wilcox_test(
    formula = diversity_shannon ~ Visite, 
    paired = TRUE,
    p.adjust.method = "BH")
```

### Beta diversity
```{r}
dist_matrix <- distance(PS.V1.V2, method = "bray")
pcoa <- cmdscale(dist_matrix)
Samp <- data.frame(sample_data(PS.V1.V2))
Samp <- merge(Samp, pcoa, by = 0)
row.names(Samp) <- Samp$Row.names
Samp$Row.names <- NULL

centroids <- aggregate(cbind(V1, V2) ~ Visite, Samp, mean)

fig.2.V1.V2.comparison.beta.diversity <- ggplot(Samp, aes(x = V1, y = V2)) +
  stat_ellipse(
    aes(color = Visite, fill = Visite),
    geom = "polygon",
    alpha = 0.1,
    level = 0.95,
    linetype = 1
  ) +
  geom_point(aes(color = Visite), size = 3, alpha = 0.7) +
  geom_point(
    data = centroids,
    aes(fill = Visite),
    size = 7,         
    shape = 21,       
    color = "black"   #
  ) +
  scale_color_manual("Visite",
    values = c("baseline 0" = "#FFD700",
  "baseline 1" = "#E69F00")
  ) +
  scale_fill_manual("Visite",
    values = c("baseline 0" = "#FFD700",
  "baseline 1" = "#E69F00")
  ) +
  xlab("Bray-Curtis distance V1") +
  ylab("Bray-Curtis distance V2") +
  boxplot.theme.2 +
  theme(
    axis.title.x = element_text(size = 24, color = "black"),  # restore x-axis title
    axis.line = element_line(color = "black"),
    legend.position = "top" )

fig.2.V1.V2.comparison.beta.diversity
#ggsave("fig.2.V1.V2.comparison.beta.diversity.pdf", plot = fig.2.V1.V2.comparison.beta.diversity, width = 5, height = 5)

######################
###### statistics ####
#####################

# Calculating bray_curtis_dist
bray_curtis_dist <- distance(PS.V1.V2, method = "bray")
Samp <- data.frame(sample_data(PS.V1.V2))

per_bray <- adonis(bray_curtis_dist ~ Visite + Geschlecht + proband_id + koerpergewicht, data = Samp,  strata = Samp$proband_id)
per_bray$aov.tab
```

